---
title: "YF time to vaccination"
author: "Carlos Haring"
date: "2026-01-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}

library(tidyverse)
library(lubridate)
library(MMWRweek)  # for MMWRweek2Date
library(gtsummary)
library(flextable)
library(summarytools)
library(fitdistrplus)

```

# Load

```{r}

# adjust path as needed
data <- read.csv("https://github.com/CarlosZHaring/Yellow-Fever-Outbreaks-and-Reactive-Vaccination-in-Africa-2010-2025-A-Scoping-Review-and-Analysis/blob/1abc1db03e3621203b42285708e203264740d86c/YF_outbreaks.csv",
                 na.strings = c("", "NA", "NaN", "N/A", "null", "NULL", "z"),
                 stringsAsFactors = FALSE)

```

# Cleaning and column creation

```{r}

# 1) Fix countries
data$country[data$country == "Cote d'voire"] <- "Cote d'Ivoire"

# 2) Parse known date columns (add any others you have)
date_cols <- c(
"first.sus.case.date","index.presentation", "index.onset", "first.conf.case.date",
  "last.conf.case.date","first.probable.case.date", "first.presumptive.positive.case.date", "last.susp.case.date","last.probable.case.date",
  "react.vac.local.start","react.vac.mass.start",
  "sus.case.reported","conf.case.reported","prob.case.reported",
  "who.outbreak.notif","who.event.notif",
  "sus.case.reported.who","conf.case.reported.who",
  "outbreak.dec","first.case.conf.nonpasteur","first.case.conf.pasteur","outbreak.inv.initiated", "activated.resp.group", "who.outbreak.notif", "who.event.notif", "outbreak.dec"
)

date_cols <- intersect(date_cols, names(data))

for (nm in date_cols) {
  # your raw looked like m/d/Y; change format if yours differs
  data[[nm]] <- as.Date(data[[nm]], "%m/%d/%Y")
}

str(data[date_cols])
```

## First of different events

```{r}
# ==== First vaccination campaign (effective earliest of local/mass) ====

data <- data %>%
  mutate(
    # ==== First vaccination campaign (effective earliest of local/mass) ====
    react.vac.eff.date = pmin(react.vac.local.start, react.vac.mass.start, na.rm = TRUE),
    # ==== First Testing  (dakar vs non) ====
    conf.test.eff.date = pmin(first.case.conf.nonpasteur, first.case.conf.pasteur, na.rm = TRUE),
    # ==== First WHO notification  (event vs outbreak vs sus vs conf) ====
    who.notif.eff.date = pmin(who.outbreak.notif, sus.case.reported.who, conf.case.reported.who, na.rm = TRUE),
    # ==== First WHO notification  (event vs outbreak vs sus vs conf) ====
    event.report.eff.date = pmin(outbreak.dec, first.case.conf.pasteur, who.outbreak.notif, conf.case.reported.who, first.case.conf.nonpasteur, na.rm = TRUE),
        # ==== last case date (event vs outbreak vs sus vs conf) ====
   last.case.eff.date = pmax(last.conf.case.date, last.probable.case.date, last.susp.case.date,
                                 na.rm = TRUE),
        # ==== first effective case date ====
     first.eff.case.date = pmin(first.sus.case.date, first.probable.case.date, first.conf.case.date, 
                                 na.rm = TRUE)
  )


```

## segeregate based on post angola and pre angola

```{r}
#"2016-07-12" is the date of the laste confirmed case in the DRC/Angola outbreak
data = data %>%
  mutate(pre.2016.outbreak = case_when(react.vac.mass.start > "2017-02-14" ~ "Post Angola/DRC",
                                     react.vac.local.start > "2017-02-14" ~ "Post Angola/DRC",
                                     react.vac.mass.start <= "2017-02-14" ~ "Pre Angola/DRC",
                                      react.vac.local.start <= "2017-02-14" ~ "Pre Angola/DRC",
                                     ))
```


## Confirmed Index Cases
This is only for index cases that were documented to be confirmed 
based on index onset, if index onset data does not exist then index presentation is used as of 1/27/26, all data is index onset. 
```{r}
data <- data %>%
  mutate(
    conf.index.case = case_when(
      index.sus.conf == "confirmed" & !is.na(index.onset) ~ index.onset,
      index.sus.conf == "confirmed" & is.na(index.onset) & !is.na(index.presentation) ~ index.presentation,
      TRUE ~ NA
    )
  )

```

## All Index Cases
index cases that were either suspected or confirmed 
```{r}
data <- data %>%
  mutate(
    all.index.case = case_when(
    !is.na(index.presentation) ~ index.presentation,
     is.na(index.presentation) & !is.na(index.onset) ~ index.onset,
      TRUE ~ NA
    )
  )
```


## Differences in weeks and days

```{r}
# ---- (whole weeks, rounded down) ----
weeks_between <- function(start_date, end_date) {
  ifelse(is.na(start_date) | is.na(end_date),
         NA_integer_,
         floor(as.numeric(end_date - start_date) / 7))
}

days_between <- function(start_date, end_date) {
  ifelse(is.na(start_date) | is.na(end_date),
         NA_integer_,
         floor(as.numeric(end_date - start_date)))
}

# Short aliases
sus   <- data$first.sus.case.date
conf  <- data$first.conf.case.date
effc  <- data$first.eff.case.date          
who   <- data$who.notif.eff.date
ctest <- data$conf.test.eff.date
erep  <- data$event.report.eff.date
allind <- data$all.index.case
confind <- data$conf.index.case 

rv_l  <- data$react.vac.local.start
rv_m  <- data$react.vac.mass.start
rv_e  <- data$react.vac.eff.date
data$effcase.to.effvac.week  <- weeks_between(effc, rv_e)


last_sus  <- data$last.susp.case
last_conf <- data$last.conf.case

# =========================================================

# effective case to vaccinations
data$effcase.to.effvac.week  <- weeks_between(effc, rv_e)
data$effcase.to.local.week   <- weeks_between(effc, rv_l)
data$effcase.to.mass.week    <- weeks_between(effc, rv_m)
data$effcase.to.effvac.days  <- days_between(effc, rv_e)
data$effcase.to.mass.days    <- days_between(effc, rv_m)
data$effcase.to.local.days    <- days_between(effc, rv_l)

# first confirmed case to vaccinations
data$confcase.to.effvac.week  <- weeks_between(conf, rv_e)
data$confcase.to.local.week   <- weeks_between(conf, rv_l)
data$confcase.to.mass.week    <- weeks_between(conf, rv_m)
data$confcase.to.effvac.days  <- days_between(conf, rv_e)
data$confcase.to.mass.days    <- days_between(conf, rv_m)
data$confcase.to.local.days    <- days_between(conf, rv_l)

#  confirmed index case to vaccinations
data$conf.index.to.effvac.week <- weeks_between(confind, rv_e)
data$conf.index.to.local.week <- weeks_between(confind, rv_l)
data$conf.index.to.mass.week <- weeks_between(confind, rv_m)
data$conf.index.to.effvac.days <- days_between(confind, rv_e)
data$conf.index.to.local.days <- days_between(confind, rv_l)
data$conf.index.to.mass.days <- days_between(confind, rv_m)


#index case (sus or confirmed) to vaccinations
data$all.index.to.effvac.week <- weeks_between(allind, rv_e)
data$all.index.to.local.week <- weeks_between(allind, rv_l)
data$all.index.to.mass.week <- weeks_between(allind, rv_m)
data$all.index.to.effvac.days <- days_between(allind, rv_e)
data$all.index.to.local.days <- days_between(allind, rv_l)
data$all.index.to.mass.days <- days_between(allind, rv_m)

# to last cases (mirroring sus/conf)
#data$effcase.to.lastsus.week  <- weeks_between(effc, last_sus)
#data$effcase.to.lastconf.week <- weeks_between(effc, last_conf)

# sus to vaccination
data$sus.effvac.days  <- days_between(data$sus.case.reported, rv_e)

# =========================================================
# B) WHO notification vs first cases & vaccinations
# =========================================================

# first cases → WHO notif
data$sus.to.who.week   <- weeks_between(sus,  who)
data$conf.to.who.week  <- weeks_between(conf, who)
data$effcase.to.who.week <- weeks_between(effc, who)
data$effcase.to.who.week <- weeks_between(effc, who)
# WHO notif → vaccinations
data$who.to.effvac.week <- weeks_between(who, rv_e)
data$who.to.local.week  <- weeks_between(who, rv_l)
data$who.to.mass.week   <- weeks_between(who, rv_m)


# Outbreak declaration to vaccination 
# =========================================================
# C) Confirmatory testing (conf.test.eff.date) vs first cases & vaccinations
# =========================================================

# first cases → confirmatory test
data$sus.to.ctest.week   <- weeks_between(sus,  ctest)
data$conf.to.ctest.week  <- weeks_between(conf, ctest)
data$effcase.to.ctest.week <- weeks_between(effc, ctest)

# confirmatory test → vaccinations
data$ctest.to.effvac.week <- weeks_between(ctest, rv_e)
data$ctest.to.local.week  <- weeks_between(ctest, rv_l)
data$ctest.to.mass.week   <- weeks_between(ctest, rv_m)

# =========================================================
# D) Event report (event.report.eff.date) vs first cases & vaccinations
# =========================================================

# first cases → event report
data$sus.to.erep.week    <- weeks_between(sus,  erep)
data$conf.to.erep.week   <- weeks_between(conf, erep)
data$effcase.to.erep.week <- weeks_between(effc, erep)

# event report → vaccinations
data$erep.to.effvac.days <- days_between(erep, rv_e)
data$erep.to.local.week  <- weeks_between(erep, rv_l)
data$erep.to.mass.week   <- weeks_between(erep, rv_m)

```




## Reactive vaccination parameter of interest 

```{r}
# function
summary_stats <- function(data, cols) {
  data.frame(
    variable = c("Index case to mass vaccination campaign", 
                 "Confirmed index case to mass vaccination campaign", 
                 "Index case to local vaccination campaign", 
                 "Confirmed index case to local vaccination campaign",
                 "Index case to first vaccination campaign",
                 "Confirmed case to first vaccination campaign",
                 "First confirmed case to mass vaccination campaign",
                 "First confirmed case to local vaccination campaign",
                 "First confirmed case to first vaccination campaign"),
    
    Median = sapply(cols, \(x) median(data[[x]], na.rm = TRUE)),
    Q1 = sapply(cols, \(x) quantile(data[[x]], 0.25, na.rm = TRUE)),
    Q3 = sapply(cols, \(x) quantile(data[[x]], 0.75, na.rm = TRUE)),
    count = sapply(cols, \(x) sum(!is.na(data[[x]])))
  )
}
# summary statistics 
summary<-summary_stats(data, c("all.index.to.mass.days", "conf.index.to.mass.days", "all.index.to.local.days", "conf.index.to.local.days", "all.index.to.effvac.days", "conf.index.to.effvac.days", "confcase.to.mass.days", "confcase.to.local.days", "confcase.to.effvac.days"))
summary


```
The 299 day max seen here is the Uganda 2019 outbreak, "delayed disbursement of funds from the government to the district slowed implementation" - EYE 2019-2020 report.

removing the uganda outbreak, does not change the estimations too much
```{r}
d1 = data %>% filter(all.index.to.mass.days	< 299)
summary_2<-summary_stats(d1, c("all.index.to.mass.days", "conf.index.to.mass.days", "all.index.to.local.days", "conf.index.to.local.days", "all.index.to.effvac.days", "conf.index.to.effvac.days", "confcase.to.mass.days", "confcase.to.local.days", "confcase.to.effvac.days"))
max(d1$all.index.to.mass.days)
min(d1$all.index.to.mass.days)
summary_2
```


# summary table

```{r}
library(flextable)
library(webshot2)
library(dplyr)

out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"
png_file <- file.path(out_dir, "summary_2.png")

ft_summary_2 <- summary_2 %>%
  tibble::rownames_to_column("Variable") %>%
  flextable() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all") %>%
  colformat_num(digits = 1) %>%
  set_caption("Summary of Time Intervals (Days)")

# Save directly as PNG
save_as_image(
  ft_summary_2,
  path = png_file,
  expand = 10,     # whitespace padding
  zoom = 2         # resolution (increase to 3 if needed)
)

ft_summary_2
```



```{r}
# =========================================================
# Create Summary Table by Pre/Post Angola Outbreak
# =========================================================

library(gtsummary)
library(dplyr)
library(flextable)
library(webshot2)

# Calculate additional variables needed
data <- data %>%
  mutate(
    # Calculate CFRs (as percentages for display)
    cfr_total = ifelse(tot.cases.reported > 0, 
                      (tot.deaths.reported / tot.cases.reported) * 100, 
                      NA_real_),
    cfr_confirmed = ifelse(tot.cases.confirmed > 0, 
                          (tot.deaths.confirmed / tot.cases.confirmed) * 100, 
                          NA_real_),
    # Calculate suspected deaths
    tot.deaths.suspected = tot.deaths.reported - tot.deaths.confirmed
  )

# Create the summary table
summary_table <- data %>%
  filter(!is.na(pre.2016.outbreak)) %>%
  dplyr::select(
    pre.2016.outbreak,
    country,
    outbreak.year,
    tot.cases.reported,
    tot.cases.confirmed,
    tot.deaths.reported,
    tot.deaths.confirmed,
    tot.deaths.suspected,
    cfr_total,
    cfr_confirmed,
    all.index.to.effvac.days,
    all.index.to.local.days,
    all.index.to.mass.days,
    confcase.to.effvac.days,
    conf.index.to.effvac.days
  ) %>%
  tbl_summary(
    by = pre.2016.outbreak,
    statistic = list(
      all_continuous() ~ "{median} ({min}, {max})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 0,
      c(cfr_total, cfr_confirmed) ~ 1
    ),
    type = list(
      country ~ "categorical",
      outbreak.year ~ "continuous",
      tot.cases.reported ~ "continuous",
      tot.cases.confirmed ~ "continuous",
      tot.deaths.reported ~ "continuous",
      tot.deaths.confirmed ~ "continuous",
      tot.deaths.suspected ~ "continuous",
      cfr_total ~ "continuous",
      cfr_confirmed ~ "continuous",
      all.index.to.effvac.days ~ "continuous",
      confcase.to.effvac.days ~ "continuous",
      conf.index.to.effvac.days ~ "continuous"
    ),
    label = list(
      country ~ "Country",
      outbreak.year ~ "Outbreak Year",
      tot.cases.reported ~ "Total Cases Reported",
      tot.cases.confirmed ~ "Total Cases Confirmed",
      tot.deaths.reported ~ "Total Deaths Reported",
      tot.deaths.confirmed ~ "Total Deaths Confirmed",
      tot.deaths.suspected ~ "Total Deaths Suspected",
      cfr_total ~ "CFR - Total Cases (%)",
      cfr_confirmed ~ "CFR - Confirmed Cases (%)",
      all.index.to.effvac.days ~ "Days: Index Case → Vaccination",
      all.index.to.local.days ~ "Days: Index Case → Local Vaccination",
      all.index.to.mass.days~ "Days: Index Case → Mass Vaccination",
      confcase.to.effvac.days ~ "Days: First Confirmed Case → Vaccination",
 
      conf.index.to.effvac.days ~ "Days: Confirmed Index Case → Vaccination"
    ),
    missing_text = "Unknown"
  ) %>%
  add_n() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Yellow Fever Outbreaks: Pre vs Post Angola/DRC**") %>%
  modify_footnote(all_stat_cols() ~ "Median (Min, Max); n (%)")

# Print the table
summary_table

# Convert to flextable and save
ft <- summary_table %>%
  as_flex_table() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever /YF/fig"

png_file <- file.path(out_dir, "outbreak_summary_table.png")
save_as_image(
  ft,
  path = png_file,
  expand = 10,
  zoom = 2
)

word_file <- file.path(out_dir, "outbreak_summary_table_min_max.docx")
save_as_docx(ft, path = word_file)


sum(data$tot.deaths.reported, na.rm = TRUE)
sum(data$tot.cases.reported, na.rm = TRUE)
sum(data$tot.deaths.reported, na.rm = TRUE)/sum(data$tot.cases.reported, na.rm = TRUE)

sum(data$tot.deaths.reported[data$pre.2016.outbreak	 == "Post Angola/DRC"], na.rm = TRUE)

sum(data$tot.cases.reported[data$pre.2016.outbreak	 == "Post Angola/DRC"], na.rm = TRUE)

sum(data$tot.deaths.reported[data$pre.2016.outbreak	 == "Post Angola/DRC"], na.rm = TRUE) /
sum(data$tot.cases.reported[data$pre.2016.outbreak	 == "Post Angola/DRC"], na.rm = TRUE)

sum(data$tot.deaths.reported[data$pre.2016.outbreak	 == "Pre Angola/DRC"], na.rm = TRUE)

sum(data$tot.cases.reported[data$pre.2016.outbreak	 == "Pre Angola/DRC"], na.rm = TRUE)

sum(data$tot.deaths.reported[data$pre.2016.outbreak	 == "Pre Angola/DRC"], na.rm = TRUE) /
sum(data$tot.cases.reported[data$pre.2016.outbreak	 == "Pre Angola/DRC"], na.rm = TRUE)


```
``` {r}
# =========================================================
# Summary Table by Pre/Post Angola Outbreak without outliers
# + use Median (Q1, Q3) instead of Median (Min, Max)
# =========================================================

library(gtsummary)
library(dplyr)
library(flextable)
library(webshot2)

# Calculate additional variables needed
data <- data %>%
  mutate(
    # Calculate CFRs (as percentages for display)
    cfr_total = ifelse(tot.cases.reported > 0,
                      (tot.deaths.reported / tot.cases.reported) * 100,
                      NA_real_),
    cfr_confirmed = ifelse(tot.cases.confirmed > 0,
                          (tot.deaths.confirmed / tot.cases.confirmed) * 100,
                          NA_real_),
    # Calculate suspected deaths
    tot.deaths.suspected = tot.deaths.reported - tot.deaths.confirmed
  )

# Create the summary table
summary_table <- data %>%
  filter(!is.na(pre.2016.outbreak)) %>%
  dplyr::select(
    pre.2016.outbreak,
    country,
    outbreak.year,
    tot.cases.reported,
    tot.cases.confirmed,
    tot.deaths.reported,
    tot.deaths.confirmed,
    tot.deaths.suspected,
    cfr_total,
    cfr_confirmed,
    all.index.to.effvac.days,
    all.index.to.local.days,
    all.index.to.mass.days,
    confcase.to.effvac.days,
    conf.index.to.effvac.days
  ) %>%
  tbl_summary(
    by = pre.2016.outbreak,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 0,
      c(cfr_total, cfr_confirmed) ~ 1
    ),
    type = list(
      country ~ "categorical",
      outbreak.year ~ "continuous",
      tot.cases.reported ~ "continuous",
      tot.cases.confirmed ~ "continuous",
      tot.deaths.reported ~ "continuous",
      tot.deaths.confirmed ~ "continuous",
      tot.deaths.suspected ~ "continuous",
      cfr_total ~ "continuous",
      cfr_confirmed ~ "continuous",
      all.index.to.effvac.days ~ "continuous",
      all.index.to.local.days ~ "continuous",
      all.index.to.mass.days ~ "continuous",
      confcase.to.effvac.days ~ "continuous",
      conf.index.to.effvac.days ~ "continuous"
    ),
    label = list(
      country ~ "Country",
      outbreak.year ~ "Outbreak Year",
      tot.cases.reported ~ "Total Cases Reported",
      tot.cases.confirmed ~ "Total Cases Confirmed",
      tot.deaths.reported ~ "Total Deaths Reported",
      tot.deaths.confirmed ~ "Total Deaths Confirmed",
      tot.deaths.suspected ~ "Total Deaths Suspected",
      cfr_total ~ "CFR - Total Cases (%)",
      cfr_confirmed ~ "CFR - Confirmed Cases (%)",
      all.index.to.effvac.days ~ "Days: Index Case → Vaccination",
      all.index.to.local.days ~ "Days: Index Case → Local Vaccination",
      all.index.to.mass.days ~ "Days: Index Case → Mass Vaccination",
      confcase.to.effvac.days ~ "Days: First Confirmed Case → Vaccination",
      conf.index.to.effvac.days ~ "Days: Confirmed Index Case → Vaccination"
    ),
    missing_text = "Unknown"
  ) %>%
  add_n() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Yellow Fever Outbreaks: Pre vs Post Angola/DRC**") %>%
  modify_footnote(all_stat_cols() ~ "Median (Q1, Q3); n (%)")

# Print the table
summary_table

# Convert to flextable and save
ft <- summary_table %>%
  as_flex_table() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"

png_file <- file.path(out_dir, "outbreak_summary_table.png")
save_as_image(
  ft,
  path = png_file,
  expand = 10,
  zoom = 2
)

word_file <- file.path(out_dir, "outbreak_summary_table_outliers_q1_q3.docx")
save_as_docx(ft, path = word_file)


# --- Your totals (unchanged) ---
sum(data$tot.deaths.reported, na.rm = TRUE)
sum(data$tot.cases.reported, na.rm = TRUE)
sum(data$tot.deaths.reported, na.rm = TRUE)/sum(data$tot.cases.reported, na.rm = TRUE)

sum(data$tot.deaths.reported[data$pre.2016.outbreak == "Post Angola/DRC"], na.rm = TRUE)
sum(data$tot.cases.reported[data$pre.2016.outbreak == "Post Angola/DRC"], na.rm = TRUE)
sum(data$tot.deaths.reported[data$pre.2016.outbreak == "Post Angola/DRC"], na.rm = TRUE) /
  sum(data$tot.cases.reported[data$pre.2016.outbreak == "Post Angola/DRC"], na.rm = TRUE)

sum(data$tot.deaths.reported[data$pre.2016.outbreak == "Pre Angola/DRC"], na.rm = TRUE)
sum(data$tot.cases.reported[data$pre.2016.outbreak == "Pre Angola/DRC"], na.rm = TRUE)
sum(data$tot.deaths.reported[data$pre.2016.outbreak == "Pre Angola/DRC"], na.rm = TRUE) /
  sum(data$tot.cases.reported[data$pre.2016.outbreak == "Pre Angola/DRC"], na.rm = TRUE)
data%>%
  filter(is.na(pre.2016.outbreak))
```

#outlier removal
```{r}
 data = data %>%
    filter(outbreak.indicator != "Uganda_2019") %>%
  filter(outbreak.indicator != "South_Sudan_2020") %>%
  filter(outbreak.indicator != "Cameroon_2021")
```

``` {r}
# =========================================================
# Summary Table by Pre/Post Angola Outbreak without outliers
# + use Median (Q1, Q3) instead of Median (Min, Max)
# =========================================================

library(gtsummary)
library(dplyr)
library(flextable)
library(webshot2)

# Calculate additional variables needed
data <- data %>%
  mutate(
    # Calculate CFRs (as percentages for display)
    cfr_total = ifelse(tot.cases.reported > 0,
                      (tot.deaths.reported / tot.cases.reported) * 100,
                      NA_real_),
    cfr_confirmed = ifelse(tot.cases.confirmed > 0,
                          (tot.deaths.confirmed / tot.cases.confirmed) * 100,
                          NA_real_),
    # Calculate suspected deaths
    tot.deaths.suspected = tot.deaths.reported - tot.deaths.confirmed
  )

# Create the summary table
summary_table <- data %>%
  filter(!is.na(pre.2016.outbreak)) %>%
  dplyr::select(
    pre.2016.outbreak,
    country,
    outbreak.year,
    tot.cases.reported,
    tot.cases.confirmed,
    tot.deaths.reported,
    tot.deaths.confirmed,
    tot.deaths.suspected,
    cfr_total,
    cfr_confirmed,
    all.index.to.effvac.days,
    all.index.to.local.days,
    all.index.to.mass.days,
    confcase.to.effvac.days,
    conf.index.to.effvac.days
  ) %>%
  # ---- EXCLUDE day-values > 299 (set to NA so they are not summarized) ----
  mutate(across(
    c(all.index.to.effvac.days,
      all.index.to.local.days,
      all.index.to.mass.days,
      confcase.to.effvac.days,
      conf.index.to.effvac.days),
    ~ ifelse(. > 299, NA_real_, .)
  )) %>%
  tbl_summary(
    by = pre.2016.outbreak,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 0,
      c(cfr_total, cfr_confirmed) ~ 1
    ),
    type = list(
      country ~ "categorical",
      outbreak.year ~ "continuous",
      tot.cases.reported ~ "continuous",
      tot.cases.confirmed ~ "continuous",
      tot.deaths.reported ~ "continuous",
      tot.deaths.confirmed ~ "continuous",
      tot.deaths.suspected ~ "continuous",
      cfr_total ~ "continuous",
      cfr_confirmed ~ "continuous",
      all.index.to.effvac.days ~ "continuous",
      all.index.to.local.days ~ "continuous",
      all.index.to.mass.days ~ "continuous",
      confcase.to.effvac.days ~ "continuous",
      conf.index.to.effvac.days ~ "continuous"
    ),
    label = list(
      country ~ "Country",
      outbreak.year ~ "Outbreak Year",
      tot.cases.reported ~ "Total Cases Reported",
      tot.cases.confirmed ~ "Total Cases Confirmed",
      tot.deaths.reported ~ "Total Deaths Reported",
      tot.deaths.confirmed ~ "Total Deaths Confirmed",
      tot.deaths.suspected ~ "Total Deaths Suspected",
      cfr_total ~ "CFR - Total Cases (%)",
      cfr_confirmed ~ "CFR - Confirmed Cases (%)",
      all.index.to.effvac.days ~ "Days: Index Case → Vaccination",
      all.index.to.local.days ~ "Days: Index Case → Local Vaccination",
      all.index.to.mass.days ~ "Days: Index Case → Mass Vaccination",
      confcase.to.effvac.days ~ "Days: First Confirmed Case → Vaccination",
      conf.index.to.effvac.days ~ "Days: Confirmed Index Case → Vaccination"
    ),
    missing_text = "Unknown"
  ) %>%
  add_n() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Yellow Fever Outbreaks: Pre vs Post Angola/DRC**") %>%
  modify_footnote(all_stat_cols() ~ "Median (Q1, Q3); n (%)")

# Print the table
summary_table

# Convert to flextable and save
ft <- summary_table %>%
  as_flex_table() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"

png_file <- file.path(out_dir, "outbreak_summary_table.png")
save_as_image(
  ft,
  path = png_file,
  expand = 10,
  zoom = 2
)

word_file <- file.path(out_dir, "outbreak_summary_table_no_outliers.docx")
save_as_docx(ft, path = word_file)
```

```{r}
# =========================================================
# Create Summary Table 
# =========================================================

library(gtsummary)
library(dplyr)
library(flextable)
library(webshot2)

# Calculate additional variables needed
data <- data %>%
  mutate(
    # Calculate CFRs (as percentages for display)
    cfr_total = ifelse(tot.cases.reported > 0, 
                      (tot.deaths.reported / tot.cases.reported) * 100, 
                      NA_real_),
    cfr_confirmed = ifelse(tot.cases.confirmed > 0, 
                          (tot.deaths.confirmed / tot.cases.confirmed) * 100, 
                          NA_real_),
    # Calculate suspected deaths
    tot.deaths.suspected = tot.deaths.reported - tot.deaths.confirmed
  )

# Create the summary table
summary_table <- data %>%
  filter(!is.na(pre.2016.outbreak)) %>%
  dplyr::select(
    pre.2016.outbreak,
    country,
    outbreak.year,
    tot.cases.reported,
    tot.cases.confirmed,
    tot.deaths.reported,
    tot.deaths.confirmed,
    tot.deaths.suspected,
    cfr_total,
    cfr_confirmed,
    all.index.to.effvac.days,
    confcase.to.effvac.days,
    conf.index.to.effvac.days
  ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{median} ({min}, {max})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 0,
      c(cfr_total, cfr_confirmed) ~ 1
    ),
    type = list(
      country ~ "categorical",
      outbreak.year ~ "continuous",
      tot.cases.reported ~ "continuous",
      tot.cases.confirmed ~ "continuous",
      tot.deaths.reported ~ "continuous",
      tot.deaths.confirmed ~ "continuous",
      tot.deaths.suspected ~ "continuous",
      cfr_total ~ "continuous",
      cfr_confirmed ~ "continuous",
      all.index.to.effvac.days ~ "continuous",
      confcase.to.effvac.days ~ "continuous",
      conf.index.to.effvac.days ~ "continuous"
    ),
    label = list(
      country ~ "Country",
      outbreak.year ~ "Outbreak Year",
      tot.cases.reported ~ "Total Cases Reported",
      tot.cases.confirmed ~ "Total Cases Confirmed",
      tot.deaths.reported ~ "Total Deaths Reported",
      tot.deaths.confirmed ~ "Total Deaths Confirmed",
      tot.deaths.suspected ~ "Total Deaths Suspected",
      cfr_total ~ "CFR - Total Cases (%)",
      cfr_confirmed ~ "CFR - Confirmed Cases (%)",
      all.index.to.effvac.days ~ "Days: Index Case → Vaccination",
      confcase.to.effvac.days ~ "Days: First Confirmed Case → Vaccination",
      conf.index.to.effvac.days ~ "Days: Confirmed Index Case → Vaccination"
    ),
    missing_text = "Unknown"
  ) %>%
  add_n() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Yellow Fever Outbreaks: Pre vs Post Angola/DRC**") %>%
  modify_footnote(all_stat_cols() ~ "Median (Min, Max); n (%)")

# Print the table
summary_table

# Convert to flextable and save
ft <- summary_table %>%
  as_flex_table() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"

png_file <- file.path(out_dir, "outbreak_summary_table.png")
save_as_image(
  ft,
  path = png_file,
  expand = 10,
  zoom = 2
)

word_file <- file.path(out_dir, "outbreak_summary_table_not_divided_min_max.docx")
save_as_docx(ft, path = word_file)
```









# reactive vaccination from index case


```{r}
library(dplyr)
library(flextable)
library(webshot2)
library(officer)

data$all.index.to.effvac.days
# Filter to rows with non-missing days-to-mass and year
pub_table <- data %>%
  arrange((all.index.case)) %>%
  filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year)) %>%
  transmute(
    Year = as.character(as.integer(outbreak.year)),     # force integer -> character (no commas)
    Country = country,
    `Days: Index → Reactive Vaccination` = all.index.to.effvac.days,
    `Total Cases` = tot.cases.reported,
    `Total Deaths` = tot.deaths.reported,
    `Confirmed Cases` = tot.cases.confirmed,
    `Confirmed Deaths` = tot.deaths.confirmed,
    CFR = tot.deaths.reported /  tot.cases.reported
  ) 
pub_table$CFR

# Build display table (convert numerics to plain character, no comma formatting)
display_tbl <- pub_table %>%
  mutate(
    `Days: Index → Reactive Vaccination` = as.character(`Days: Index → Reactive Vaccination`),
    `Total Cases` = ifelse(is.na(`Total Cases`), "Unknown", as.character(as.integer(`Total Cases`))),
    `Total Deaths` = ifelse(is.na(`Total Deaths`), "Unknown", as.character(as.integer(`Total Deaths`))),
    `Confirmed Cases` = ifelse(is.na(`Confirmed Cases`), "Unknown", as.character(as.integer(`Confirmed Cases`))),
    `Confirmed Deaths` = ifelse(is.na(`Confirmed Deaths`), "Unknown", as.character(as.integer(`Confirmed Deaths`))),
    `CFR` = ifelse(is.na(`Total Deaths`), "Unknown", as.character(round(CFR,2)))
  )

ft_pub <- flextable(display_tbl) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(j = 1, align = "left", part = "body") %>%
  align(j = 2:ncol(display_tbl), align = "center", part = "all") %>%
  autofit() %>%
  fontsize(size = 10, part = "all") %>%
  set_header_labels(
    Year = "Year",
    Country = "Country",
    `Days: Index → Reactive Vaccination` = "Days: Index Case to Reactive Vaccination",
    `Total Cases` = "Total Cases",
    `Total Deaths` = "Total Deaths",
    `Confirmed Cases` = "Confirmed Cases",
    `Confirmed Deaths` = "Confirmed Deaths",
    `CFR` = "CFR"
  )
pub_table$`Days: Index → Reactive Vaccination`

days_num_valid <- pub_table$`Days: Index → Reactive Vaccination`
days_num_valid <- days_num_valid[!is.na(days_num_valid)]

footer_txt <- sprintf(
  "Days: mean = %.1f; median = %.1f; mode = %s; IQR = %.1f–%.1f; range = %.1f–%.1f; n = %d complete outbreaks",
  mean(days_num_valid),
  median(days_num_valid),
  names(sort(table(days_num_valid), decreasing = TRUE))[1],
  quantile(days_num_valid, 0.25),
  quantile(days_num_valid, 0.75),
  min(days_num_valid),
  max(days_num_valid),
  length(days_num_valid)
)

ft_pub <- ft_pub %>%
  add_footer_lines(values = footer_txt) %>%
  align(align = "left", part = "footer")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
png_file <- file.path(out_dir, "table_1_no234_all.png")
docx_file <- file.path(out_dir, "table_1_no234_all.docx")

try(save_as_image(ft_pub, path = png_file, expand = 8, zoom = 2), silent = TRUE)
try(save_as_docx(ft_pub, path = docx_file), silent = TRUE)

ft_pub


```


```{r}
pub_table <- data %>%
  filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year))

pub_table %>%
  group_by(pre.2016.outbreak) %>%
  summarise(n= n(), per = n()/nrow(pub_table) * 100)
```

##wilcoxon test

```{r}
coin::wilcox_test(data = data, all.index.to.effvac.days ~ as.factor(pre.2016.outbreak), distribution = "exact") # use this for ties
all_a <- data %>%
    filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year)) %>%
  filter(pre.2016.outbreak == "Post Angola/DRC")

all_b <- data %>%
    filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year)) %>%
  filter(pre.2016.outbreak == "Pre Angola/DRC")

wilcox.test(all_a$all.index.to.mass.days, all_b$all.index.to.mass.days)
```

# CFR and death anaylsis 
```{r}
row = nrow(data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)))
data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n(),perc = round(n/row*100,1))

data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n())
```

```{r}
data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n(),
            average_cases = mean(tot.cases.reported),
            lower_bound=quantile(tot.cases.reported, .25),
upper_bound=quantile(tot.cases.reported, .75),
median = median(tot.cases.reported))

data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n(),
            average_cases = mean(tot.cases.reported),
            lower_bound=quantile(tot.cases.reported, .25),
upper_bound=quantile(tot.cases.reported, .75),
median = median(tot.cases.reported))
```


### wilcoxon of median reported cases
```{r}
library(coin)
all_a <- data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported))

all_a$pre.2016.outbreak = as.factor(all_a$pre.2016.outbreak)

wilcox_test(tot.cases.reported ~ pre.2016.outbreak, data = all_a,
            distribution = "exact")
```


```{r}
f1=data %>%
 filter(!is.na(tot.cases.reported)) %>%
 filter(!is.na(tot.deaths.reported))

```

```{r}
tot = nrow(data %>%
    filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)))

var_ratio_by_period <- data %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)) %>%
  summarise(
    n = n(),
    perc = n()/tot *100,
    Xbar = mean(tot.cases.reported),

    r_hat = sum(tot.deaths.reported) / sum(tot.cases.reported),

    s2_X = var(tot.cases.reported),
    s2_Y = var(tot.deaths.reported),
    s_XY = cov(tot.deaths.reported, tot.cases.reported),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
  )
var_ratio_by_period
```

```{r}
tot = nrow(data %>%
    filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)))

var_ratio_by_period <- data %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(
    n = n(),
    perc = n()/tot *100,
    Xbar = mean(tot.cases.reported),

    r_hat = sum(tot.deaths.reported) / sum(tot.cases.reported),

    s2_X = var(tot.cases.reported),
    s2_Y = var(tot.deaths.reported),
    s_XY = cov(tot.deaths.reported, tot.cases.reported),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
  )
var_ratio_by_period
```

```{r}

t_score=(var_ratio_by_period[1,5]-var_ratio_by_period[2,5])/sqrt(var_ratio_by_period[1,10]+var_ratio_by_period[2,10])
t_score
n = var_ratio_by_period[1,2 ]+ var_ratio_by_period[2,2]-2

2 * pt(-abs(as.numeric(t_score)), df = as.numeric(n))
```


# plots

```{r}
Theme = function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(size = base_size + 4, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = base_size + 2),
      axis.title = element_text(size = base_size + 2, face = "bold"),
      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),
       legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```

# histogram of delay
```{r}
data$tot.cases.reported
data_1 = data %>%
    filter(outbreak.indicator != "Uganda_2019") %>%
  filter(outbreak.indicator != "South_Sudan_2020") %>%
  filter(outbreak.indicator != "Cameroon_2024")

ggplot(data = data_1[data_1$all.index.to.effvac.days < 234,], 
       aes(x = as.Date(all.index.case), y = all.index.to.effvac.days)
       ) + 
  geom_point(width = 75, 
           alpha = 1,
           fill = "blue",
           colour = "black",
           linewidth = .5, aes(
           size = tot.cases.reported)
          ) +
  geom_vline(
    aes(xintercept = as.Date("2017-02-14"), colour = "End of 2016 DRC outbreak"),
    linetype = "solid",
    linewidth = 1
) +
scale_colour_manual(
  name = NULL,
  values = c("End of 2016 DRC outbreak" = "red")
)  +
  scale_size_continuous(name = "Total Reported Cases",
                        breaks = c(10, 100, 500, 1000),
                        range = c(3, 15)
                        
                      ) +
  
   scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
    ) +
  xlab("Date of Index Case"
  ) +
  ylab("Index case to Reactive Vaccination (Days)"
  ) +
   annotate("label", 
           label = "Angola 2015-2016", 
           x = as.Date("2015-12-5"), 
           y = 80
           ) +
     annotate("label", 
           label = "Kenya 2022", 
           x = as.Date("2022-1-5"), 
           y = 210
           ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))
  ) + 
  labs() +
  Theme() +
  theme(legend.position = "bottom")
  
ggsave(
  filename = "figure_1.pdf",
  plot = last_plot(),
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = fig_path
)

ggsave(
  filename = "figure_1.png",
  plot = last_plot(),
  device = ragg::agg_png,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  path = fig_path
)
```

```{r}
Theme_new <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(
        size = base_size + 8,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      plot.subtitle = element_text(size = base_size + 2),

      axis.title = element_text(size = base_size + 6, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),

      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),

      legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```


# histogram of cases log
```{r}
ggplot(data = data[!is.na(data$tot.cases.reported) & !is.na(data$tot.deaths.reported), ], 
       aes(x = as.Date(all.index.case), y = tot.deaths.reported/tot.cases.reported, size = tot.cases.reported)
       ) + 
  geom_point(
           fill = "blue",
           colour = "black"
          
          ) +
  geom_vline(
    aes(xintercept = as.Date("2017-02-14"), colour = "End of Angola/DRC outbreak"),
    linetype = "solid",
    linewidth = 1
) +
scale_colour_manual(
  name = NULL,
  values = c("End of Angola/DRC outbreak" = "red")
) + 
   scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
    )  +
  xlab("Date of Index Case"
  ) +
  ylab("Case Fatality Ratio"
  ) +
  coord_cartesian(ylim = c(0,1)
  ) + 
  scale_size_continuous(name = "Total Reported Cases",
                         breaks = c(10, 100, 500, 1000),
                        range = c(2, 13)) +
  Theme_new() +
  theme(legend.position = "bottom")
  data[data$all.index.to.effvac.days < 234 & 
       (!is.na(data$all.index.case) & !is.na(data$tot.cases.reported)) & !is.na(data$tot.deaths.reported), ]
  
ggsave(
  filename = "figure_3.pdf",
  plot = last_plot(),
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = fig_path
)

ggsave(
  filename = "figure_3.png",
  plot = last_plot() +Theme_new(base_size = 12),
  device = ragg::agg_png,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  path = fig_path
)
```


```{r}
data %>%
  dplyr::select(outbreak.indicator,cfr_total) %>%
  arrange(desc(cfr_total))
data %>%
  dplyr::select(outbreak.indicator,tot.cases.reported, cfr_total) %>%
  filter(!is.na(cfr_total)) %>%
  arrange(desc(tot.cases.reported))
```


#