---
title: "YF time to vaccination"
author: "Carlos Haring"
date: "2026-01-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}

library(tidyverse)
library(lubridate)
library(MMWRweek)  # for MMWRweek2Date
library(gtsummary)
library(flextable)
library(summarytools)
library(fitdistrplus)
library(webshot2)
library(coin)
```

# Load and create output directory

```{r}
out_dir <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/fig"
data <- read.csv("~/Documents/Documents/Yellow Fever/YF_scoping_review/data/YF_outbreaks.csv",
                 na.strings = c("", "NA", "NaN", "N/A", "null", "NULL", "z"),
                 stringsAsFactors = FALSE)
```

# Cleaning and column creation

## Country fixing, data column creation
```{r}

# Fix countries
data$country[data$country == "Cote d'voire"] <- "Cote d'Ivoire"

# Parse date columns
date_cols <- c(
"index.presentation", "index.onset", "react.vac.local.start", "react.vac.mass.start",
"first.case confirmed", "who.notif", "outbreak.dec"
)

date_cols <- intersect(date_cols, names(data))

for (nm in date_cols) {
  data[[nm]] <- as.Date(data[[nm]], "%m/%d/%Y")
}

str(data[date_cols])
```

## First of different events
```{r}
# ==== First vaccination campaign (effective earliest of local/mass) ====
data <- data %>%
  mutate(
    # ==== First vaccination campaign (effective earliest of local/mass) ====
    react.vac.eff.date = pmin(react.vac.local.start, react.vac.mass.start, na.rm = TRUE))
```

## segeregate based on post angola and pre angola

```{r}
#"2016-07-12" is the date of the last confirmed case in the DRC/Angola outbreak
data = data %>%
  mutate(pre.2016.outbreak = case_when(react.vac.mass.start > "2017-02-14" ~ "Post Angola/DRC",
                                     react.vac.local.start > "2017-02-14" ~ "Post Angola/DRC",
                                     react.vac.mass.start <= "2017-02-14" ~ "Pre Angola/DRC",
                                      react.vac.local.start <= "2017-02-14" ~ "Pre Angola/DRC",
                                     ))
```


## Confirmed Index Cases
This is only for index cases that were documented to be confirmed based on index onset, if index onset data does not exist then index presentation is used instead.
```{r}
data <- data %>%
  mutate(
    conf.index.case = case_when(
      index.sus.conf == "confirmed" & !is.na(index.onset) ~ index.onset,
      index.sus.conf == "confirmed" & is.na(index.onset) & !is.na(index.presentation) ~ index.presentation,
      TRUE ~ NA
    )
  )

```

## All Index Cases

index cases that were either suspected or confirmed or probable
```{r}
data <- data %>%
  mutate(
    all.index.case = case_when(
    !is.na(index.presentation) ~ index.presentation,
     is.na(index.presentation) & !is.na(index.onset) ~ index.onset,
      TRUE ~ NA
    )
  )
```


## Differences in weeks and days function
```{r}
# Helper function
days_between <- function(start_date, end_date) {
  ifelse(is.na(start_date) | is.na(end_date),
         NA_integer_,
         floor(as.numeric(end_date - start_date)))
}

# Short alias
allind <- data$all.index.case
rv_e  <- data$react.vac.eff.date


# index case (sus or confirmed) to vaccinations
data$all.index.to.effvac.days <- days_between(allind, rv_e)
```

# number of outbreaks
```{r}
nrow(data)
```

# number of outbreaks pre vs post angola
```{r}
data %>%
  group_by(pre.2016.outbreak) %>%
  summarise(perc =round(n()/nrow(data)*100,2), n= n())
```

# creating mortality and time to reactive vaccination dataframes
```{r}
ttv_df = data %>%
  filter(!is.na(all.index.to.effvac.days))

mort_df = data %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported))
```

# Time to reactive vaccination analysis

## quick histogram
```{r}
hist(ttv_df$all.index.to.effvac.days)
```

## examine outliers 
```{r}
ttv_df %>%
  filter(all.index.to.effvac.days >= 200) %>%
  dplyr::select(outbreak.identifier, outbreak.year, country, 
                all.index.to.effvac.days)
```
### South Sudan 2020 rational for exclusion
"Implementation [of South Sudan reactive vaccination campaign] was delayed until
October 2020 because of COVID-19" -Weekly epidemiological record, Global yellow 
fever update, 2020.

### Uganda 2019 rational for exlusion
"The ICG supported an [Uganda] outbreak
response targeting 1 665 402 people in August 2020,
after delays due partly to COVID-19" -Weekly epidemiological record, Global 
yellow fever update, 2020.

### Cameroon 2021 rational for exlusion
The epidemiological evidence between the linkage of the index case in Cameroon 
and the vaccination campaign is to small relative to the large delay seen.


## outlier removal
```{r}
 ttv_df = ttv_df %>%
    filter(outbreak.identifier != "Uganda_2019") %>%
  filter(outbreak.identifier != "South_Sudan_2020") %>%
  filter(outbreak.identifier != "Cameroon_2021")
```

## percentage and number pre and post angola campaign
```{r}
ttv_df %>%
  group_by(pre.2016.outbreak) %>%
  summarise(n= n(), per = n()/nrow(pub_table) * 100)
```

## Table 1, table of reactive vaccination campaigns 
```{r}
# Filter to rows with non-missing days-to-mass and year
pub_table <- ttv_df %>%
  arrange((all.index.case)) %>%
  transmute(
    Year = as.character(as.integer(outbreak.year)),     # force integer -> character (no commas)
    Country = country,
    `Days: Index → Reactive Vaccination` = all.index.to.effvac.days,
    `Total Cases` = tot.cases.reported,
    `Total Deaths` = tot.deaths.reported,
    CFR = tot.deaths.reported / tot.cases.reported
  ) 
# Build display table (convert numerics to plain character, no comma formatting)
display_tbl <- pub_table %>%
  mutate(
    `Days: Index → Reactive Vaccination` = as.character(`Days: Index → Reactive Vaccination`),
    `Total Cases` = ifelse(is.na(`Total Cases`), "Unknown", as.character(as.integer(`Total Cases`))),
    `Total Deaths` = ifelse(is.na(`Total Deaths`), "Unknown", as.character(as.integer(`Total Deaths`))),
    `CFR` = ifelse(is.na(CFR), "Unknown", as.character(round(CFR,2)))
  )

ft_pub <- flextable(display_tbl) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(j = 1, align = "left", part = "body") %>%
  align(j = 2:ncol(display_tbl), align = "center", part = "all") %>%
  autofit() %>%
  fontsize(size = 10, part = "all") %>%
  set_header_labels(
    Year = "Year",
    Country = "Country",
    `Days: Index → Reactive Vaccination` = "Days: Index Case to Reactive Vaccination",
    `Total Cases` = "Total Cases",
    `Total Deaths` = "Total Deaths",
    `Confirmed Cases` = "Confirmed Cases",
    `Confirmed Deaths` = "Confirmed Deaths",
    `CFR` = "CFR"
  )

days_num_valid <- pub_table$`Days: Index → Reactive Vaccination`
days_num_valid <- days_num_valid[!is.na(days_num_valid)]

footer_txt <- sprintf(
  "Days: mean = %.1f; median = %.1f; mode = %s; IQR = %.1f–%.1f; range = %.1f–%.1f; n = %d complete outbreaks",
  mean(days_num_valid),
  median(days_num_valid),
  names(sort(table(days_num_valid), decreasing = TRUE))[1],
  quantile(days_num_valid, 0.25),
  quantile(days_num_valid, 0.75),
  min(days_num_valid),
  max(days_num_valid),
  length(days_num_valid)
)

ft_pub <- ft_pub %>%
  add_footer_lines(values = footer_txt) %>%
  align(align = "left", part = "footer")

# Save outputs
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
png_file <- file.path(out_dir, "table_1.png")
docx_file <- file.path(out_dir, "table_1.docx")

try(save_as_image(ft_pub, path = png_file, expand = 8, zoom = 2), silent = TRUE)
try(save_as_docx(ft_pub, path = docx_file), silent = TRUE)

ft_pub


```

## wilcoxon test
```{r}
coin::wilcox_test(data = ttv_df, all.index.to.effvac.days ~ as.factor(pre.2016.outbreak), distribution = "exact") # use this for ties
```
## summary table pre and post angola
```{r}
med_delay = ttv_df %>%
  group_by(as.factor(pre.2016.outbreak)) %>%
    summarise(
            lower_bound=quantile(all.index.to.effvac.days, .25),
upper_bound=quantile(all.index.to.effvac.days, .75),
median = median(all.index.to.effvac.days))
med_delay
pre_delay_med=as.numeric(med_delay$median[2])
post_delay_med=as.numeric(med_delay$median[1])
```

## smallest and largest delays 

### minimum
```{r}
ttv_df %>%
  arrange(all.index.to.effvac.days) %>%
   dplyr::select(outbreak.identifier, all.index.to.effvac.days) %>%
  filter(all.index.to.effvac.days == min(all.index.to.effvac.days))

```

### maximum
```{r}
ttv_df %>%
  arrange(all.index.to.effvac.days) %>%
   dplyr::select(outbreak.identifier, all.index.to.effvac.days) %>%
  filter(all.index.to.effvac.days == max(all.index.to.effvac.days))
```
## Theme for figure 1 plot
```{r}
Theme = function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(size = base_size + 4, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = base_size + 2),
      axis.title = element_text(size = base_size + 2, face = "bold"),
      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),
       legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size-2,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```


## Figure 1: Delay of reactive vaccinations
```{r}
cutoff_date <- as.Date("2017-02-14")

ggplot(
  data = ttv_df,
  aes(x = as.Date(all.index.case), y = all.index.to.effvac.days)
) + 
  geom_point(
    aes(size = tot.cases.reported),
    alpha = 1,
    fill = "blue",
    colour = "black",
    linewidth = .5
  ) +
  
  geom_vline(
    aes(xintercept = cutoff_date, colour = "End of 2016 Angola/DRC outbreak"),
    linetype = "solid",
    linewidth = 1
  ) +
  
  geom_segment(
    aes(
      x = min(as.Date(ttv_df$all.index.case), na.rm = TRUE),
      xend = cutoff_date,
      y = pre_delay_med,
      yend = pre_delay_med,
      colour = "Median RVC delay (days)"
    ),
    linetype = "dashed",
    linewidth = 1
  ) +
  
  geom_segment(
    aes(
      x = cutoff_date,
      xend = max(as.Date(ttv_df$all.index.case), na.rm = TRUE),
      y = post_delay_med,
      yend = post_delay_med,
      colour = "Median RVC delay (days)"
    ),
    linetype = "dashed",
    linewidth = 1
  ) +
  
  scale_colour_manual(
    name = NULL,
    values = c(
      "End of 2016 Angola/DRC outbreak" = "blue",
      "Median RVC delay (days)" = "red"
    )
  ) +
  
  scale_size_continuous(
    name = "Total Reported Cases",
    breaks = c(10, 100, 500, 1000),
    range = c(3, 15)
  ) +
  
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  
  xlab("Date of Index Case") +
  ylab("Index case to Reactive Vaccination (Days)") +
  
  annotate("label",
           label = "Angola 2015-2016",
           x = as.Date("2015-12-05"),
           y = 80) +
  annotate("label",
           label = "Kenya 2022",
           x = as.Date("2022-01-05"),
           y = 210) +
  
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))) +
  theme(legend.position = "bottom") +
    theme(
  legend.position = "bottom",
  legend.direction = "horizontal"
) +
guides(
  colour = guide_legend(nrow = 2, byrow = TRUE),
  size = guide_legend(nrow = 2, byrow = TRUE)
)
  
ggsave(
  filename = "figure_1.pdf",
  plot = last_plot(),
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = out_dir
)

ggsave(
  filename = "figure_1.png",
  plot = last_plot(),
  device = ragg::agg_png,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  path = out_dir
)

ggsave(
  filename = "figure_1.tiff",
  plot = last_plot() +
    theme(
      plot.background   = element_rect(fill = "white", color = NA),
      panel.background  = element_rect(fill = "white", color = NA),
      panel.grid.major  = element_line(color = "grey80", linewidth = 0.4),
      panel.grid.minor  = element_line(color = "grey90", linewidth = 0.2)
    ),
  device = ragg::agg_tiff,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 900,
  path = out_dir,
  bg = "white",
  compression = "lzw"
)
```


# CFR and death anaylsis 

### row count, basic summary by period and overall of CFR
```{r}
row_count= nrow(mort_df)
mort_df %>%
  group_by(pre.2016.outbreak) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported) *100, n = n(), perc = round(n/row_count*100,1))

mort_df %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported) * 100, n = n())
```
```{r}
row_count = nrow(mort_df)

var_ratio_by_period <- mort_df %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)) %>%
  summarise(
    n = n(),
    Xbar = mean(tot.cases.reported),

    r_hat = sum(tot.deaths.reported) / sum(tot.cases.reported),

    s2_X = var(tot.cases.reported),
    s2_Y = var(tot.deaths.reported),
    s_XY = cov(tot.deaths.reported, tot.cases.reported),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
  )
var_ratio_by_period
```


## variance and SE calculation for CFR
```{r}
row_count = nrow(mort_df)

var_ratio_by_period <- mort_df %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(
    n = n(),
    perc = round(n/row_count *100,2),
    Xbar = mean(tot.cases.reported),

    r_hat = sum(tot.deaths.reported) / sum(tot.cases.reported),

    s2_X = var(tot.cases.reported),
    s2_Y = var(tot.deaths.reported),
    s_XY = cov(tot.deaths.reported, tot.cases.reported),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
  )
var_ratio_by_period

mort_pre_median = as.numeric(var_ratio_by_period$r_hat[2])
mort_post_median = as.numeric(var_ratio_by_period$r_hat[1])
```

## Welch's T-test for CFR
```{r}

t_score_a = (var_ratio_by_period[1,5] - var_ratio_by_period[2,5]) / sqrt(var_ratio_by_period[1,9] + var_ratio_by_period[2,9])
welch_df = ((var_ratio_by_period[1,9]  + var_ratio_by_period[2,9]) ^ 2) / ((var_ratio_by_period[1,9] ^ 2 / (var_ratio_by_period[1,2]-1)) + (var_ratio_by_period[2,9] ^ 2 / (var_ratio_by_period[2,2]-1)))
2 * pt(-abs(as.numeric(t_score_a)), df = as.numeric(welch_df))
```

## smallest and largest CFRs 

### minimums
```{r}
mort_df %>%
  mutate(CFR = tot.deaths.reported/tot.cases.reported) %>%
  filter(CFR == min(CFR)) %>%
  dplyr::select(CFR, outbreak.identifier)

```

### maximums
```{r}
mort_df %>%
  mutate(CFR = tot.deaths.reported/tot.cases.reported) %>%
  filter(CFR == max(CFR)) %>%
  dplyr::select(CFR, outbreak.identifier)

```

## Median reported deaths overall and grouped
```{r}
mort_period_median = mort_df %>%
  group_by(pre.2016.outbreak) %>%
  summarise(lower_bound=quantile(tot.cases.reported, .25),
upper_bound=quantile(tot.cases.reported, .75),
median = median(tot.cases.reported))
mort_period_median


mort_df %>%
  summarise(average_cases = mean(tot.cases.reported),
            lower_bound=quantile(tot.cases.reported, .25),
upper_bound=quantile(tot.cases.reported, .75),
median = median(tot.cases.reported))
```
## wilcoxon of median reported cases
```{r}
wilcox_test(tot.cases.reported ~ as.factor(pre.2016.outbreak), data = mort_df,
            distribution = "exact")
```


## smallest and largest outbreaks by total cases 

### minimums
```{r}
mort_df %>%
  filter(tot.cases.reported == min(tot.cases.reported)) %>%
  dplyr::select(tot.cases.reported, outbreak.identifier)

```

### maximums
```{r}
mort_df %>%
   filter(tot.cases.reported == max(tot.cases.reported)) %>%
   dplyr::select(tot.cases.reported, outbreak.identifier)

```


## figure 3  theme
```{r}
Theme_new <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(
        size = base_size + 8,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      plot.subtitle = element_text(size = base_size + 2),

      axis.title = element_text(size = base_size + 6, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),

      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),

      legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size-2,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```


# Figure 2 
```{r}
ggplot(data = mort_df, 
            aes(x = as.Date(all.index.case),
                y = tot.deaths.reported / tot.cases.reported,
                size = tot.cases.reported)) + 
  geom_point(
    fill = "blue",
    colour = "black"
  ) +
  geom_vline(
    aes(xintercept = as.Date("2017-02-14"),
        colour = "End of 2016 Angola/DRC outbreak"),
    linetype = "solid",
    linewidth = 1
  ) +
geom_segment(
  aes(
    x = min(as.Date(mort_df$all.index.case), na.rm = TRUE),
    xend = as.Date("2017-02-14"),
    y = mort_pre_median,
    yend = mort_pre_median,
    colour = "Overall CFR"
  ),
  linetype = "dashed",
  linewidth = 1
) +
geom_segment(
  aes(
    x = as.Date("2017-02-14"),
    xend = max(as.Date(mort_df$all.index.case), na.rm = TRUE),
    y = mort_post_median,
    yend = mort_post_median,
    colour = "Overall CFR"
  ),
  linetype = "dashed",
  linewidth = 1
) +
scale_colour_manual(
  name = NULL,
  values = c(
    "End of 2016 Angola/DRC outbreak" = "blue",
    "Overall CFR" = "red"
  )) + 
  scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
  ) +
  xlab("Date of Index Case") +
  ylab("Case Fatality Ratio") +
  coord_cartesian(ylim = c(0, 1)) + 
  scale_size_continuous(
    name = "Total Reported Cases",
    breaks = c(10, 100, 500, 1000),
    range = c(2, 13)
  ) +
  Theme_new() +
  theme(
    legend.position = "bottom",
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA)
  ) +
theme(
  legend.position = "bottom",
  legend.direction = "horizontal"
) +
guides(
  size = guide_legend(order = 1),
  colour = guide_legend(order = 2, nrow = 2, byrow = TRUE)
)
  data[data$all.index.to.effvac.days < 234 & 
       (!is.na(data$all.index.case) & !is.na(data$tot.cases.reported)) & !is.na(data$tot.deaths.reported), ]
  
ggsave(
  filename = "figure_2.pdf",
  plot = last_plot(),
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = out_dir
)

ggsave(
  filename = "figure_2.png",
  plot = last_plot() +Theme_new(base_size = 12),
  device = ragg::agg_png,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  path = out_dir
)

ggsave(
  filename = "figure_2.tiff",
  plot = last_plot() +
    theme(
    panel.grid.major = element_line(color = "grey80"),
    panel.grid.minor = element_line(color = "grey90"),
      plot.background  = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA)
    ),
  device = ragg::agg_tiff,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 900,
  path = out_dir,
  bg = "white",
  compression = "lzw"
)
```
