---
title: "YF time to vaccination"
author: "Carlos Haring"
date: "2026-01-21"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Setup

```{r}

library(tidyverse)
library(lubridate)
library(MMWRweek)  # for MMWRweek2Date
library(gtsummary)
library(flextable)
library(summarytools)
library(fitdistrplus)
library(webshot2)
```

# Load

```{r}
data <- read.csv("https://raw.githubusercontent.com/CarlosZHaring/Yellow-Fever-Outbreaks-and-Reactive-Vaccination-in-Africa-2010-2025-A-Scoping-Review-and-Analysis/refs/heads/main/data/YF_outbreaks.csv?token=GHSAT0AAAAAAEA6O6QF6HU55MJUTKNE3VC42R5RCSQ",
                 na.strings = c("", "NA", "NaN", "N/A", "null", "NULL", "z"),
                 stringsAsFactors = FALSE)
```

# Cleaning and column creation

## Country fixing, data column creation
```{r}

# Fix countries
data$country[data$country == "Cote d'voire"] <- "Cote d'Ivoire"

# Parse date columns
date_cols <- c(
"index.presentation", "index.onset", "react.vac.local.start", "react.vac.mass.start",
"first.case confirmed", "who.notif", "outbreak.dec"
)

date_cols <- intersect(date_cols, names(data))

for (nm in date_cols) {
  data[[nm]] <- as.Date(data[[nm]], "%m/%d/%Y")
}

str(data[date_cols])
```

## First of different events
```{r}
# ==== First vaccination campaign (effective earliest of local/mass) ====
data <- data %>%
  mutate(
    # ==== First vaccination campaign (effective earliest of local/mass) ====
    react.vac.eff.date = pmin(react.vac.local.start, react.vac.mass.start, na.rm = TRUE))
```

## segeregate based on post angola and pre angola

```{r}
#"2016-07-12" is the date of the last confirmed case in the DRC/Angola outbreak
data = data %>%
  mutate(pre.2016.outbreak = case_when(react.vac.mass.start > "2017-02-14" ~ "Post Angola/DRC",
                                     react.vac.local.start > "2017-02-14" ~ "Post Angola/DRC",
                                     react.vac.mass.start <= "2017-02-14" ~ "Pre Angola/DRC",
                                      react.vac.local.start <= "2017-02-14" ~ "Pre Angola/DRC",
                                     ))
```


## Confirmed Index Cases
This is only for index cases that were documented to be confirmed based on index onset, if index onset data does not exist then index presentation is used instead.
```{r}
data <- data %>%
  mutate(
    conf.index.case = case_when(
      index.sus.conf == "confirmed" & !is.na(index.onset) ~ index.onset,
      index.sus.conf == "confirmed" & is.na(index.onset) & !is.na(index.presentation) ~ index.presentation,
      TRUE ~ NA
    )
  )

```

## All Index Cases

index cases that were either suspected or confirmed or probable
```{r}
data <- data %>%
  mutate(
    all.index.case = case_when(
    !is.na(index.presentation) ~ index.presentation,
     is.na(index.presentation) & !is.na(index.onset) ~ index.onset,
      TRUE ~ NA
    )
  )
```


## Differences in weeks and days function
```{r}
# Helper function
days_between <- function(start_date, end_date) {
  ifelse(is.na(start_date) | is.na(end_date),
         NA_integer_,
         floor(as.numeric(end_date - start_date)))
}

# Short alias
allind <- data$all.index.case
rv_e  <- data$react.vac.eff.date


# index case (sus or confirmed) to vaccinations
data$all.index.to.effvac.days <- days_between(allind, rv_e)
```


# creating mortality and time to reactive vaccination dataframes
```{r}
ttv_df = data %>%
  filter(!is.na(all.index.to.effvac.days))

mort_df = data %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported))
```

# Time to reactive vaccination analysis

## quick histogram
```{r}
hist(ttv_df$all.index.to.effvac.days)
```

## examine outliers 
```{r}
ttv_df %>%
  filter(all.index.to.effvac.days >= 200) %>%
  dplyr::select(outbreak.identifier, outbreak.year, country, 
                all.index.to.effvac.days)
```
### South Sudan 2020 rational for exclusion
"Implementation [of South Sudan reactive vaccination campaign] was delayed until
October 2020 because of COVID-19" -Weekly epidemiological record, Global yellow 
fever update, 2020.

### Uganda 2019 rational for exlusion
"The ICG supported an [Uganda] outbreak
response targeting 1 665 402 people in August 2020,
after delays due partly to COVID-19" -Weekly epidemiological record, Global 
yellow fever update, 2020.

### Cameroon 2021 rational for exlusion
The epidemiological evidence between the linkage of the index case in Cameroon 
and the vaccination campaign is to small relative to the large delay seen.


#outlier removal
```{r}
 data = data %>%
    filter(outbreak.indicator != "Uganda_2019") %>%
  filter(outbreak.indicator != "South_Sudan_2020") %>%
  filter(outbreak.indicator != "Cameroon_2021")
```

``` {r}
# =========================================================
# Summary Table by Pre/Post Angola Outbreak without outliers
# + use Median (Q1, Q3) instead of Median (Min, Max)
# =========================================================

library(gtsummary)
library(dplyr)
library(flextable)
library(webshot2)

# Calculate additional variables needed
data <- data %>%
  mutate(
    # Calculate CFRs (as percentages for display)
    cfr_total = ifelse(tot.cases.reported > 0,
                      (tot.deaths.reported / tot.cases.reported) * 100,
                      NA_real_),
    cfr_confirmed = ifelse(tot.cases.confirmed > 0,
                          (tot.deaths.confirmed / tot.cases.confirmed) * 100,
                          NA_real_),
    # Calculate suspected deaths
    tot.deaths.suspected = tot.deaths.reported - tot.deaths.confirmed
  )

# Create the summary table
summary_table <- data %>%
  filter(!is.na(pre.2016.outbreak)) %>%
  dplyr::select(
    pre.2016.outbreak,
    country,
    outbreak.year,
    tot.cases.reported,
    tot.cases.confirmed,
    tot.deaths.reported,
    tot.deaths.confirmed,
    tot.deaths.suspected,
    cfr_total,
    cfr_confirmed,
    all.index.to.effvac.days,
    all.index.to.local.days,
    all.index.to.mass.days,
    confcase.to.effvac.days,
    conf.index.to.effvac.days
  ) %>%
  # ---- EXCLUDE day-values > 299 (set to NA so they are not summarized) ----
  mutate(across(
    c(all.index.to.effvac.days,
      all.index.to.local.days,
      all.index.to.mass.days,
      confcase.to.effvac.days,
      conf.index.to.effvac.days),
    ~ ifelse(. > 299, NA_real_, .)
  )) %>%
  tbl_summary(
    by = pre.2016.outbreak,
    statistic = list(
      all_continuous() ~ "{median} ({p25}, {p75})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 0,
      c(cfr_total, cfr_confirmed) ~ 1
    ),
    type = list(
      country ~ "categorical",
      outbreak.year ~ "continuous",
      tot.cases.reported ~ "continuous",
      tot.cases.confirmed ~ "continuous",
      tot.deaths.reported ~ "continuous",
      tot.deaths.confirmed ~ "continuous",
      tot.deaths.suspected ~ "continuous",
      cfr_total ~ "continuous",
      cfr_confirmed ~ "continuous",
      all.index.to.effvac.days ~ "continuous",
      all.index.to.local.days ~ "continuous",
      all.index.to.mass.days ~ "continuous",
      confcase.to.effvac.days ~ "continuous",
      conf.index.to.effvac.days ~ "continuous"
    ),
    label = list(
      country ~ "Country",
      outbreak.year ~ "Outbreak Year",
      tot.cases.reported ~ "Total Cases Reported",
      tot.cases.confirmed ~ "Total Cases Confirmed",
      tot.deaths.reported ~ "Total Deaths Reported",
      tot.deaths.confirmed ~ "Total Deaths Confirmed",
      tot.deaths.suspected ~ "Total Deaths Suspected",
      cfr_total ~ "CFR - Total Cases (%)",
      cfr_confirmed ~ "CFR - Confirmed Cases (%)",
      all.index.to.effvac.days ~ "Days: Index Case → Vaccination",
      all.index.to.local.days ~ "Days: Index Case → Local Vaccination",
      all.index.to.mass.days ~ "Days: Index Case → Mass Vaccination",
      confcase.to.effvac.days ~ "Days: First Confirmed Case → Vaccination",
      conf.index.to.effvac.days ~ "Days: Confirmed Index Case → Vaccination"
    ),
    missing_text = "Unknown"
  ) %>%
  add_n() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Yellow Fever Outbreaks: Pre vs Post Angola/DRC**") %>%
  modify_footnote(all_stat_cols() ~ "Median (Q1, Q3); n (%)")

# Print the table
summary_table

# Convert to flextable and save
ft <- summary_table %>%
  as_flex_table() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"

png_file <- file.path(out_dir, "outbreak_summary_table.png")
save_as_image(
  ft,
  path = png_file,
  expand = 10,
  zoom = 2
)

word_file <- file.path(out_dir, "outbreak_summary_table_no_outliers.docx")
save_as_docx(ft, path = word_file)
```

```{r}
# =========================================================
# Create Summary Table 
# =========================================================

library(gtsummary)
library(dplyr)
library(flextable)
library(webshot2)

# Calculate additional variables needed
data <- data %>%
  mutate(
    # Calculate CFRs (as percentages for display)
    cfr_total = ifelse(tot.cases.reported > 0, 
                      (tot.deaths.reported / tot.cases.reported) * 100, 
                      NA_real_),
    cfr_confirmed = ifelse(tot.cases.confirmed > 0, 
                          (tot.deaths.confirmed / tot.cases.confirmed) * 100, 
                          NA_real_),
    # Calculate suspected deaths
    tot.deaths.suspected = tot.deaths.reported - tot.deaths.confirmed
  )

# Create the summary table
summary_table <- data %>%
  filter(!is.na(pre.2016.outbreak)) %>%
  dplyr::select(
    pre.2016.outbreak,
    country,
    outbreak.year,
    tot.cases.reported,
    tot.cases.confirmed,
    tot.deaths.reported,
    tot.deaths.confirmed,
    tot.deaths.suspected,
    cfr_total,
    cfr_confirmed,
    all.index.to.effvac.days,
    confcase.to.effvac.days,
    conf.index.to.effvac.days
  ) %>%
  tbl_summary(
    statistic = list(
      all_continuous() ~ "{median} ({min}, {max})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    digits = list(
      all_continuous() ~ 0,
      c(cfr_total, cfr_confirmed) ~ 1
    ),
    type = list(
      country ~ "categorical",
      outbreak.year ~ "continuous",
      tot.cases.reported ~ "continuous",
      tot.cases.confirmed ~ "continuous",
      tot.deaths.reported ~ "continuous",
      tot.deaths.confirmed ~ "continuous",
      tot.deaths.suspected ~ "continuous",
      cfr_total ~ "continuous",
      cfr_confirmed ~ "continuous",
      all.index.to.effvac.days ~ "continuous",
      confcase.to.effvac.days ~ "continuous",
      conf.index.to.effvac.days ~ "continuous"
    ),
    label = list(
      country ~ "Country",
      outbreak.year ~ "Outbreak Year",
      tot.cases.reported ~ "Total Cases Reported",
      tot.cases.confirmed ~ "Total Cases Confirmed",
      tot.deaths.reported ~ "Total Deaths Reported",
      tot.deaths.confirmed ~ "Total Deaths Confirmed",
      tot.deaths.suspected ~ "Total Deaths Suspected",
      cfr_total ~ "CFR - Total Cases (%)",
      cfr_confirmed ~ "CFR - Confirmed Cases (%)",
      all.index.to.effvac.days ~ "Days: Index Case → Vaccination",
      confcase.to.effvac.days ~ "Days: First Confirmed Case → Vaccination",
      conf.index.to.effvac.days ~ "Days: Confirmed Index Case → Vaccination"
    ),
    missing_text = "Unknown"
  ) %>%
  add_n() %>%
  modify_header(label = "**Characteristic**") %>%
  modify_caption("**Yellow Fever Outbreaks: Pre vs Post Angola/DRC**") %>%
  modify_footnote(all_stat_cols() ~ "Median (Min, Max); n (%)")

# Print the table
summary_table

# Convert to flextable and save
ft <- summary_table %>%
  as_flex_table() %>%
  autofit() %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(align = "center", part = "all") %>%
  align(j = 1, align = "left", part = "body") %>%
  fontsize(size = 10, part = "all") %>%
  padding(padding = 6, part = "all")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"

png_file <- file.path(out_dir, "outbreak_summary_table.png")
save_as_image(
  ft,
  path = png_file,
  expand = 10,
  zoom = 2
)

word_file <- file.path(out_dir, "outbreak_summary_table_not_divided_min_max.docx")
save_as_docx(ft, path = word_file)
```









# reactive vaccination from index case


```{r}
library(dplyr)
library(flextable)
library(webshot2)
library(officer)

data$all.index.to.effvac.days
# Filter to rows with non-missing days-to-mass and year
pub_table <- data %>%
  arrange((all.index.case)) %>%
  filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year)) %>%
  transmute(
    Year = as.character(as.integer(outbreak.year)),     # force integer -> character (no commas)
    Country = country,
    `Days: Index → Reactive Vaccination` = all.index.to.effvac.days,
    `Total Cases` = tot.cases.reported,
    `Total Deaths` = tot.deaths.reported,
    `Confirmed Cases` = tot.cases.confirmed,
    `Confirmed Deaths` = tot.deaths.confirmed,
    CFR = tot.deaths.reported /  tot.cases.reported
  ) 
pub_table$CFR

# Build display table (convert numerics to plain character, no comma formatting)
display_tbl <- pub_table %>%
  mutate(
    `Days: Index → Reactive Vaccination` = as.character(`Days: Index → Reactive Vaccination`),
    `Total Cases` = ifelse(is.na(`Total Cases`), "Unknown", as.character(as.integer(`Total Cases`))),
    `Total Deaths` = ifelse(is.na(`Total Deaths`), "Unknown", as.character(as.integer(`Total Deaths`))),
    `Confirmed Cases` = ifelse(is.na(`Confirmed Cases`), "Unknown", as.character(as.integer(`Confirmed Cases`))),
    `Confirmed Deaths` = ifelse(is.na(`Confirmed Deaths`), "Unknown", as.character(as.integer(`Confirmed Deaths`))),
    `CFR` = ifelse(is.na(`Total Deaths`), "Unknown", as.character(round(CFR,2)))
  )

ft_pub <- flextable(display_tbl) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(j = 1, align = "left", part = "body") %>%
  align(j = 2:ncol(display_tbl), align = "center", part = "all") %>%
  autofit() %>%
  fontsize(size = 10, part = "all") %>%
  set_header_labels(
    Year = "Year",
    Country = "Country",
    `Days: Index → Reactive Vaccination` = "Days: Index Case to Reactive Vaccination",
    `Total Cases` = "Total Cases",
    `Total Deaths` = "Total Deaths",
    `Confirmed Cases` = "Confirmed Cases",
    `Confirmed Deaths` = "Confirmed Deaths",
    `CFR` = "CFR"
  )
pub_table$`Days: Index → Reactive Vaccination`

days_num_valid <- pub_table$`Days: Index → Reactive Vaccination`
days_num_valid <- days_num_valid[!is.na(days_num_valid)]

footer_txt <- sprintf(
  "Days: mean = %.1f; median = %.1f; mode = %s; IQR = %.1f–%.1f; range = %.1f–%.1f; n = %d complete outbreaks",
  mean(days_num_valid),
  median(days_num_valid),
  names(sort(table(days_num_valid), decreasing = TRUE))[1],
  quantile(days_num_valid, 0.25),
  quantile(days_num_valid, 0.75),
  min(days_num_valid),
  max(days_num_valid),
  length(days_num_valid)
)

ft_pub <- ft_pub %>%
  add_footer_lines(values = footer_txt) %>%
  align(align = "left", part = "footer")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
png_file <- file.path(out_dir, "table_1_no234_all.png")
docx_file <- file.path(out_dir, "table_1_no234_all.docx")

try(save_as_image(ft_pub, path = png_file, expand = 8, zoom = 2), silent = TRUE)
try(save_as_docx(ft_pub, path = docx_file), silent = TRUE)

ft_pub


```


```{r}
pub_table <- data %>%
  filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year))

pub_table %>%
  group_by(pre.2016.outbreak) %>%
  summarise(n= n(), per = n()/nrow(pub_table) * 100)
```

##wilcoxon test

```{r}
coin::wilcox_test(data = data, all.index.to.effvac.days ~ as.factor(pre.2016.outbreak), distribution = "exact") # use this for ties
all_a <- data %>%
    filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year)) %>%
  filter(pre.2016.outbreak == "Post Angola/DRC")

all_b <- data %>%
    filter(!is.na(all.index.to.effvac.days), !is.na(outbreak.year)) %>%
  filter(pre.2016.outbreak == "Pre Angola/DRC")

wilcox.test(all_a$all.index.to.mass.days, all_b$all.index.to.mass.days)
```

# CFR and death anaylsis 
```{r}
row = nrow(data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)))
data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n(),perc = round(n/row*100,1))

data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n())
```

```{r}
data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n(),
            average_cases = mean(tot.cases.reported),
            lower_bound=quantile(tot.cases.reported, .25),
upper_bound=quantile(tot.cases.reported, .75),
median = median(tot.cases.reported))

data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported)) %>%
  summarise(CFR = sum(tot.deaths.reported)/sum(tot.cases.reported), n = n(),
            average_cases = mean(tot.cases.reported),
            lower_bound=quantile(tot.cases.reported, .25),
upper_bound=quantile(tot.cases.reported, .75),
median = median(tot.cases.reported))
```


### wilcoxon of median reported cases
```{r}
library(coin)
all_a <- data %>%
  filter(!is.na(tot.cases.reported)) %>%
  filter(!is.na(tot.deaths.reported))

all_a$pre.2016.outbreak = as.factor(all_a$pre.2016.outbreak)

wilcox_test(tot.cases.reported ~ pre.2016.outbreak, data = all_a,
            distribution = "exact")
```


```{r}
f1=data %>%
 filter(!is.na(tot.cases.reported)) %>%
 filter(!is.na(tot.deaths.reported))

```

```{r}
tot = nrow(data %>%
    filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)))

var_ratio_by_period <- data %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)) %>%
  summarise(
    n = n(),
    perc = n()/tot *100,
    Xbar = mean(tot.cases.reported),

    r_hat = sum(tot.deaths.reported) / sum(tot.cases.reported),

    s2_X = var(tot.cases.reported),
    s2_Y = var(tot.deaths.reported),
    s_XY = cov(tot.deaths.reported, tot.cases.reported),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
  )
var_ratio_by_period
```

```{r}
tot = nrow(data %>%
    filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)))

var_ratio_by_period <- data %>%
  filter(!is.na(tot.cases.reported), !is.na(tot.deaths.reported)) %>%
  group_by(pre.2016.outbreak) %>%
  summarise(
    n = n(),
    perc = n()/tot *100,
    Xbar = mean(tot.cases.reported),

    r_hat = sum(tot.deaths.reported) / sum(tot.cases.reported),

    s2_X = var(tot.cases.reported),
    s2_Y = var(tot.deaths.reported),
    s_XY = cov(tot.deaths.reported, tot.cases.reported),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
  )
var_ratio_by_period
```

```{r}

t_score=(var_ratio_by_period[1,5]-var_ratio_by_period[2,5])/sqrt(var_ratio_by_period[1,10]+var_ratio_by_period[2,10])
t_score
n = var_ratio_by_period[1,2 ]+ var_ratio_by_period[2,2]-2

2 * pt(-abs(as.numeric(t_score)), df = as.numeric(n))
```


# plots

```{r}
Theme = function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(size = base_size + 4, face = "bold", hjust = 0.5),
      plot.subtitle = element_text(size = base_size + 2),
      axis.title = element_text(size = base_size + 2, face = "bold"),
      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),
       legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```

# histogram of delay
```{r}
data$tot.cases.reported
data_1 = data %>%
    filter(outbreak.indicator != "Uganda_2019") %>%
  filter(outbreak.indicator != "South_Sudan_2020") %>%
  filter(outbreak.indicator != "Cameroon_2024")

ggplot(data = data_1[data_1$all.index.to.effvac.days < 234,], 
       aes(x = as.Date(all.index.case), y = all.index.to.effvac.days)
       ) + 
  geom_point(width = 75, 
           alpha = 1,
           fill = "blue",
           colour = "black",
           linewidth = .5, aes(
           size = tot.cases.reported)
          ) +
  geom_vline(
    aes(xintercept = as.Date("2017-02-14"), colour = "End of 2016 DRC outbreak"),
    linetype = "solid",
    linewidth = 1
) +
scale_colour_manual(
  name = NULL,
  values = c("End of 2016 DRC outbreak" = "red")
)  +
  scale_size_continuous(name = "Total Reported Cases",
                        breaks = c(10, 100, 500, 1000),
                        range = c(3, 15)
                        
                      ) +
  
   scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
    ) +
  xlab("Date of Index Case"
  ) +
  ylab("Index case to Reactive Vaccination (Days)"
  ) +
   annotate("label", 
           label = "Angola 2015-2016", 
           x = as.Date("2015-12-5"), 
           y = 80
           ) +
     annotate("label", 
           label = "Kenya 2022", 
           x = as.Date("2022-1-5"), 
           y = 210
           ) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.15))
  ) + 
  labs() +
  Theme() +
  theme(legend.position = "bottom")
  
ggsave(
  filename = "figure_1.pdf",
  plot = last_plot(),
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = fig_path
)

ggsave(
  filename = "figure_1.png",
  plot = last_plot(),
  device = ragg::agg_png,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  path = fig_path
)
```

```{r}
Theme_new <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(
        size = base_size + 8,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      plot.subtitle = element_text(size = base_size + 2),

      axis.title = element_text(size = base_size + 6, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),

      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),

      legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```


# histogram of cases log
```{r}
ggplot(data = data[!is.na(data$tot.cases.reported) & !is.na(data$tot.deaths.reported), ], 
       aes(x = as.Date(all.index.case), y = tot.deaths.reported/tot.cases.reported, size = tot.cases.reported)
       ) + 
  geom_point(
           fill = "blue",
           colour = "black"
          
          ) +
  geom_vline(
    aes(xintercept = as.Date("2017-02-14"), colour = "End of Angola/DRC outbreak"),
    linetype = "solid",
    linewidth = 1
) +
scale_colour_manual(
  name = NULL,
  values = c("End of Angola/DRC outbreak" = "red")
) + 
   scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
    )  +
  xlab("Date of Index Case"
  ) +
  ylab("Case Fatality Ratio"
  ) +
  coord_cartesian(ylim = c(0,1)
  ) + 
  scale_size_continuous(name = "Total Reported Cases",
                         breaks = c(10, 100, 500, 1000),
                        range = c(2, 13)) +
  Theme_new() +
  theme(legend.position = "bottom")
  data[data$all.index.to.effvac.days < 234 & 
       (!is.na(data$all.index.case) & !is.na(data$tot.cases.reported)) & !is.na(data$tot.deaths.reported), ]
  
ggsave(
  filename = "figure_3.pdf",
  plot = last_plot(),
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = fig_path
)

ggsave(
  filename = "figure_3.png",
  plot = last_plot() +Theme_new(base_size = 12),
  device = ragg::agg_png,
  width = 8.5,
  height = 5.5,
  units = "in",
  dpi = 300,
  path = fig_path
)
```


```{r}
data %>%
  dplyr::select(outbreak.indicator,cfr_total) %>%
  arrange(desc(cfr_total))
data %>%
  dplyr::select(outbreak.indicator,tot.cases.reported, cfr_total) %>%
  filter(!is.na(cfr_total)) %>%
  arrange(desc(tot.cases.reported))
```


#
