---
title: "4_13_vacc_rate"
author: "Carlos Haring"
date: "2026-04-13"
output:
  pdf_document: default
  html_document: default
---
# Loading relevant packages and files
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## library addition
```{r}
library(tidyverse)
library(lubridate)
library(MMWRweek)  # for MMWRweek2Date
library(gtsummary)
library(flextable)
library(summarytools)
library(fitdistrplus)
library(webshot2)
library(coin)

font_add("Arial", regular = "arial.ttf")
showtext_auto()
```

## paths
```{r}
out_dir <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/fig"
src_path <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/src"
data_path <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/data"

file_path <- file.path(data_path, "YF_RVCs.csv")
```

## file reading 
```{r}
df <- read_csv(file_path)
```

# Cleaning/column creation

## date creation
```{r}
# parse dates (M/D/YYYY format)
df <- df %>%
  mutate(
    start_date = mdy(react.vac.start.date.one),
    end_date   = mdy(react.vac.end.date.one)
  ) %>%
  filter(vacc.marker != "DRC_2014_1_pv") #because there is an aggregate estimate that works better
```

## establish vaccination types 
```{r}
df <- df %>%
  mutate(
    # split vacc.marker
    base_marker = str_remove(vacc.marker, "_(pv|im|admin)$"),
    type = case_when(
      str_detect(vacc.marker, "_pv$") ~ "pv",
      str_detect(vacc.marker, "_im$") ~ "im",
      str_detect(vacc.marker, "_admin$") ~ "admin",
      TRUE ~ "admin"   # treat null as admin
    ),
    
    # priority: im/pv > admin
    priority = case_when(
      type %in% c("im", "pv") ~ 1,
      TRUE ~ 2
    )
  )
```


# establish geography markers
```{r}
geo_cols <- c(
"Geographic.marker.1",
"Geographic.marker.2",
"Geographic.marker.3"
)

df <- df %>%
  unite("geo_id", all_of(geo_cols), sep = "|", na.rm = TRUE, remove = FALSE)
```


# creating
```{r}
df_clean <- df %>%
  mutate(
    reac.vac.perc.one = if_else(
      is.na(reac.vac.perc.one) & !is.na(reac.vac.one) & !is.na(reac.target.pop.one),
      reac.vac.one / reac.target.pop.one * 100,
      reac.vac.perc.one
    ),
    reac.target.pop.one = if_else(
      is.na(reac.target.pop.one) & !is.na(reac.vac.one) & !is.na(reac.vac.perc.one),
      (reac.vac.one / reac.vac.perc.one) * 100,
      reac.target.pop.one
    ),
    reac.vac.one = if_else(
      is.na(reac.vac.one) & !is.na(reac.target.pop.one) & !is.na(reac.vac.perc.one),
      (reac.vac.perc.one / 100) * reac.target.pop.one,
      reac.vac.one
    )
  )
df_clean
```

## creating and following priority rule
```{r}
df_clean <- df_clean %>%
  mutate(
    completeness = rowSums(!is.na(pick(
      reac.vac.one,
      reac.target.pop.one,
      reac.vac.perc.one
    )))
  ) %>%
  group_by(base_marker, geo_id) %>%
  arrange(desc(completeness), priority, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup()

df_clean
```

## mark by country and year
```{r}
df_clean <- df_clean %>%
  mutate(
    country = str_extract(base_marker, "^[^_]+"),
    year    = as.numeric(str_extract(base_marker, "\\d{4}"))
  )
```


## creating the dataframe of relevent vaccination information 
```{r}
campaign_df <- df_clean %>%
  group_by(base_marker) %>%
  summarise(
    outbreak.identifier = first(outbreak.identifier),
    country = first(country),
    year = first(year),
    start = min(start_date, na.rm = TRUE),
    end   = max(end_date, na.rm = TRUE),
    duration = as.numeric(end - start + 1),
        target_pop = sum(reac.target.pop.one),
    vaccinated_pop = sum(reac.vac.one),
    log_estimate = -log(1-sum(vaccinated_pop)/sum(target_pop))/duration,
    Constant_log_estimate = (1-exp(-log_estimate)),
    n_rows = n(),
    coverage = sum(reac.vac.one, na.rm = TRUE) /
                        sum(reac.target.pop.one, na.rm = TRUE) * 100,
    rate = sum(coverage) / sum(duration),
    .groups = "drop"
  )
```

## Classify pre vs post-Angola Campaigns
```{r}
cutoff <- as.Date("2017-02-14")

campaign_df <- campaign_df %>%
  mutate(
    period = if_else(start < cutoff, "pre_angola", "post_angola")
  )
```

# Results 

## the pooled weighted estimate and standard error
```{r}
ratio_var <- campaign_df %>%
  filter(!is.na(coverage), !is.na(duration)) %>%
  summarise(
    n = n(),
    Xbar = mean(duration),
    Ybar = mean(coverage),  
    
    r_hat = sum(coverage) / sum(duration),
    
    s2_X = var(duration),
    s2_Y = var(coverage),
    s_XY = cov(duration, coverage, use = "complete.obs"),
    
    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),
    
    se_ratio = sqrt(var_ratio)
  )

ratio_var
```

## the pooled weighted estimate and standard error by period
```{r}
tot = nrow(campaign_df %>%
             filter(!is.na(coverage), !is.na(duration))) 

var_ratio_by_period <- campaign_df %>%
  filter(!is.na(coverage), !is.na(duration)) %>%
  group_by(period) %>%
  summarise(
    n = n(),
    perc = round(n()/tot *100,2),
    Xbar = mean(duration),

    r_hat = sum(coverage) / sum(duration),

    s2_X = var(duration),
    s2_Y = var(coverage),
    s_XY = cov(duration, coverage),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
    .groups = "drop"
  )
var_ratio_by_period

```


## T test
```{r}
t_score_a = (var_ratio_by_period[1,5] - var_ratio_by_period[2,5]) / sqrt(var_ratio_by_period[1,10] + var_ratio_by_period[2,10])
n = tot - 2
2 * pt(-as.numeric(t_score_a), df = n)
```


# histogram of pooled reactive vaccination
```{r}
pub_table <- campaign_df %>%
  arrange((start)) %>%
  filter(!is.na(rate)) %>%
  transmute(
    `Start date` = start,
    `End date` = end, # force integer, character (no commas)
    Country = country,
    `Target Population` = round(target_pop,0),
    `Coverage (%)` = round(coverage,2),
    `Vaccination rate` = round(rate,2)
  ) 

pub_table = pub_table %>%
  relocate(Country, .before = `Start date`)

ft_pub <- flextable(pub_table) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(j = 1, align = "left", part = "body") %>%
  align(j = 2:ncol(pub_table), align = "center", part = "all") %>%
  autofit() %>%
  fontsize(size = 10, part = "all") %>%
  set_header_labels(
    Country = "Country",
    `Start date` = "Start date",
    `End date` = "End date", 
    `Target Population` = "Target Population",
    `Coverage (%)` = "Coverage (%)",
    `Vaccination rate`  = "Vaccination rate (% per day)"
  )

png_file <- file.path(out_dir, "table_2S.png")
docx_file <- file.path(out_dir, "table_2S.docx")

save_as_image(ft_pub, path = png_file, expand = 8, zoom = 2)
save_as_docx(ft_pub, path = docx_file)

ft_pub
```






