---
title: "4_13_vacc_rate"
author: "Carlos Haring"
date: "2026-04-13"
output:
  pdf_document: default
  html_document: default
---
# Loading relevant packages and files
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## library addition
```{r}
library(tidyverse)
library(lubridate)
library(MMWRweek)  # for MMWRweek2Date
library(gtsummary)
library(flextable)
library(summarytools)
library(fitdistrplus)
library(webshot2)
library(coin)

font_add("Arial", regular = "arial.ttf")
showtext_auto()
```

## paths
```{r}
out_dir <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/fig"
src_path <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/src"
data_path <- "~/Documents/Documents/Yellow Fever/YF_scoping_review/data"

file_path <- file.path(data_path, "YF_RVCs.csv")
```

## file reading 
```{r}
df <- read_csv(file_path)
```

# Cleaning/column creation

## date creation
```{r}
# parse dates (M/D/YYYY format)
df <- df %>%
  mutate(
    start_date = mdy(react.vac.start.date.one),
    end_date   = mdy(react.vac.end.date.one)
  ) %>%
  filter(vacc.marker != "DRC_2014_1_pv") #because there is an aggregate estimate that works better
```

## establish vaccination types 
```{r}
df <- df %>%
  mutate(
    # split vacc.marker
    base_marker = str_remove(vacc.marker, "_(pv|im|admin)$"),
    type = case_when(
      str_detect(vacc.marker, "_pv$") ~ "pv",
      str_detect(vacc.marker, "_im$") ~ "im",
      str_detect(vacc.marker, "_admin$") ~ "admin",
      TRUE ~ "admin"   # treat null as admin
    ),
    
    # priority: im/pv > admin
    priority = case_when(
      type %in% c("im", "pv") ~ 1,
      TRUE ~ 2
    )
  )
```


# establish geography markers
```{r}
geo_cols <- c(
"Geographic.marker.1",
"Geographic.marker.2",
"Geographic.marker.3"
)

df <- df %>%
  unite("geo_id", all_of(geo_cols), sep = "|", na.rm = TRUE, remove = FALSE)
```


# creating
```{r}
df_clean <- df %>%
  mutate(
    reac.vac.perc.one = if_else(
      is.na(reac.vac.perc.one) & !is.na(reac.vac.one) & !is.na(reac.target.pop.one),
      reac.vac.one / reac.target.pop.one * 100,
      reac.vac.perc.one
    ),
    reac.target.pop.one = if_else(
      is.na(reac.target.pop.one) & !is.na(reac.vac.one) & !is.na(reac.vac.perc.one),
      (reac.vac.one / reac.vac.perc.one) * 100,
      reac.target.pop.one
    ),
    reac.vac.one = if_else(
      is.na(reac.vac.one) & !is.na(reac.target.pop.one) & !is.na(reac.vac.perc.one),
      (reac.vac.perc.one / 100) * reac.target.pop.one,
      reac.vac.one
    )
  )
df_clean
```

## creating and following priority rule
```{r}
df_clean <- df_clean %>%
  mutate(
    completeness = rowSums(!is.na(pick(
      reac.vac.one,
      reac.target.pop.one,
      reac.vac.perc.one
    )))
  ) %>%
  group_by(base_marker, geo_id) %>%
  arrange(desc(completeness), priority, .by_group = TRUE) %>%
  slice_head(n = 1) %>%
  ungroup()

df_clean
```

## mark by country and year
```{r}
df_clean <- df_clean %>%
  mutate(
    country = str_extract(base_marker, "^[^_]+"),
    year    = as.numeric(str_extract(base_marker, "\\d{4}"))
  )
```


## creating the dataframe of relevent vaccination information 
```{r}
campaign_df <- df_clean %>%
  group_by(base_marker) %>%
  summarise(
    outbreak.identifier = first(outbreak.identifier),
    country = first(country),
    year = first(year),
    start = min(start_date, na.rm = TRUE),
    end   = max(end_date, na.rm = TRUE),
    duration = as.numeric(end - start + 1),
        target_pop = sum(reac.target.pop.one),
    vaccinated_pop = sum(reac.vac.one),
    log_estimate = -log(1-sum(vaccinated_pop)/sum(target_pop))/duration,
    Constant_log_estimate = (1-exp(-log_estimate)),
    n_rows = n(),
    coverage = sum(reac.vac.one, na.rm = TRUE) /
                        sum(reac.target.pop.one, na.rm = TRUE),
    rate = sum(coverage) / sum(duration),
    .groups = "drop"
  )
```

## classify pre vs post angola camepiagns
```{r}
cutoff <- as.Date("2017-02-14")

campaign_df <- campaign_df %>%
  mutate(
    period = if_else(start < cutoff, "pre_angola", "post_angola")
  )
```

# Results 

## the pooled weighted variance
```{r}
ratio_var <- campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  summarise(
    n = n(),
    Xbar = mean(duration, na.rm = TRUE),
    Ybar = mean(coverage_weighted, na.rm = TRUE),  
    
    r_hat = sum(coverage_weighted, na.rm = TRUE) / sum(duration, na.rm = TRUE),
    
    s2_X = var(duration, na.rm = TRUE),
    s2_Y = var(coverage_weighted, na.rm = TRUE),
    s_XY = cov(duration, coverage_weighted, use = "complete.obs"),
    
    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),
    
    se_ratio = sqrt(var_ratio)
  )

ratio_var
```

## variance weighed by period
```{r}
tot = nrow(campaign_df %>%
    filter(!is.na(coverage_weighted), !is.na(duration)))

var_ratio_by_period_a <- campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  group_by(period) %>%
  summarise(
    n = n(),
    perc = n()/tot *100,
    Xbar = mean(duration),

    r_hat = sum(coverage_weighted) / sum(duration),

    s2_X = var(duration),
    s2_Y = var(coverage_weighted),
    s_XY = cov(duration, coverage_weighted),

    var_ratio = (1 / (n * Xbar^2)) *
      (r_hat^2 * s2_X + s2_Y - 2 * r_hat * s_XY),

    se_ratio = sqrt(var_ratio),
    .groups = "drop"
  )
var_ratio_by_period_a

```


## T test
```{r}

t_score_a=(var_ratio_by_period_a[1,5]-var_ratio_by_period_a[2,5])/sqrt(var_ratio_by_period_a[1,10]+var_ratio_by_period_a[2,10])
(var_ratio_by_period_a[1,10]-var_ratio_by_period_a[2,10])
sqrt(var_ratio_by_period_a[1,10]+var_ratio_by_period_a[2,10])
t_score_a
n = tot-2
2 * pt(-as.numeric(t_score_a), df = n)
```



# coverage for the vaccination campeigns 

##overall median
```{r}
campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration))%>%
  summarise(round(median(coverage_weighted)*100,1), Q1 = round(quantile(coverage_weighted, .25)*100,1),  Q3 = round(quantile(coverage_weighted, .75)*100,1), n = n())
```


##min and max
```{r}
campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  group_by(base_marker) %>%
  summarise(coverage = round((coverage_weighted)*100,1)) %>%
  arrange(coverage)
```

##segregated median by period
```{r}
rows = nrow(campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)))

campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration))%>%
  group_by(period) %>%
  summarise( coverage = round(median(coverage_weighted)*100,2), Q1 = round(quantile(coverage_weighted, .25)*100,1),  Q3 = round(quantile(coverage_weighted, .75)*100,1), n = n(), perc = n()/rows)
```

##fishers test for these
```{r}
pre = (campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  filter(period == "pre_angola"))$coverage_weighted

post = (campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  filter(period == "post_angola"))$coverage_weighted

wilcox.test(pre,post)
```


# log weights 
```{r}
log_pooled_weighted <- campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  filter(coverage_weighted<1) %>%
  summarise(
    pooled_rate = sum(-log(1-coverage_weighted))/sum(duration)
  )
log_pooled_weighted

campaign_df %>%
  filter(!is.na(coverage_weighted), !is.na(duration)) %>%
  filter(coverage_weighted<1) %>%
  dplyr::select(outbreak.identifier, country, year, duration, coverage_weighted, rate_weighted) %>%
  group_by(outbreak.identifier) %>%
  summarise(
    pooled_rate = sum(-log(1-coverage_weighted))/sum(duration)
  )
```











## sudan outbreak metrics
```{r}
campaign_df %>%
  filter(base_marker == "SouthSudan_2024_1") %>%
  dplyr::select(base_marker, coverage_weighted, coverage_unweighted,start, end,rate_unweighted,rate_weighted,coverage_weighted,coverage_unweighted)
campaign_df
```

```{r}
setwd("/Users/carlosh/Documents/Documents/Yellow Fever/YF/src")
df = read_csv("output.csv")

colnames(df)[1] = "outbreak.identifier"
```




# grouped by outbreak
```{r}
pooled_country_outbreak <- campaign_df %>%
  group_by(base_marker) %>%
  summarise(
    pooled_rate_weighted = sum(coverage_weighted, na.rm = TRUE) / sum(duration, na.rm = TRUE),
    pooled_rate_unweighted = sum(coverage_unweighted, na.rm = TRUE) / sum(duration, na.rm = TRUE),
    n_campaigns = n(),
    start_date = min(start, na.rm = TRUE),
    end_date = max(end, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(pooled_rate_weighted != 0)

```

```{r}

library(tidyverse)
library(lubridate)
library(MMWRweek)  # for MMWRweek2Date
library(gtsummary)
library(flextable)
library(summarytools)
library(fitdistrplus)
Theme <- function(base_size = 12, base_family = "sans") {
  theme_minimal(base_size = base_size, base_family = base_family) +
    theme(
      plot.title = element_text(
        size = base_size + 8,
        face = "bold",
        hjust = 0.5,
        margin = margin(b = 8)
      ),
      plot.subtitle = element_text(size = base_size + 2),

      axis.title = element_text(size = base_size + 6, face = "bold"),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),

      axis.text = element_text(size = base_size),
      axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
      axis.text.y = element_text(size = base_size),

      legend.position = "bottom",
      legend.justification = "left",
      legend.box.just = "left",
      legend.key.width = unit(0.15, "in"),
      legend.spacing.x = unit(0.1, "in"),
      legend.spacing.y = unit(0.05, "in"),
      legend.box.spacing = unit(0.1, "in"),
      legend.margin = margin(0, 0, 0, 0),

      legend.title = element_text(
        size = base_size + 6,
        face = "bold",
        hjust = 0
      ),
      legend.text = element_text(size = base_size,
                                margin = margin(l = 2)),

      strip.text = element_text(size = base_size + 1, face = "bold"),
      panel.grid.minor = element_blank()
    )
}
```
# histogram of pooled reactive vaccination
```{r}
# campaign_df$duration_group <- cut(
#   campaign_df$duration,
#   #breaks = c(0, 5, 10, 20 ,40 ,60, Inf),
#   #labels = c("0–5", "6–10", "11–20", "21-40", "41–60", "61+"),
#   include.lowest = TRUE
# )

p = ggplot(data = campaign_df, 
       aes(x = as.Date(start), y = rate_weighted*100,
        size = as.numeric(coverage_weighted*100))
    ) + 
  geom_point(
           colour = "black",
           linewidth = .5,
           alpha = .9
          ) +
  geom_vline(
    aes(xintercept = as.Date("2017-02-14"), colour = "End of Angola/DRC outbreak"),
    linetype = "solid",
    linewidth = 1
) +
  scale_colour_manual(
  name = NULL,
  values = c("End of Angola/DRC outbreak" = "red")
) +
#     annotate("label", 
#            label = "End of Angola/DRC outbreak", 
#            x = as.Date("2017-02-14"), 
#            y = 50,
#            h_just = 0
#            )+
#   annotate("label", 
#            label = "South Sudan 
# 2020", 
#            x = as.Date("2020-10-23"), 
#            y = 50,
#            h_just = 0
#            ) +
   scale_x_date(
    date_breaks = "1 year",
    date_labels = "%Y"
    ) +
  xlab("Start date of Reactive Vaccination"
  ) +
  ylab("Target population Vaccinated per Day (%)"
  ) +
  scale_y_continuous(breaks = seq.default(from =0 ,to =25, by = 5)
  ) + 
  scale_size_continuous(
    name = "Target Population Coverage (%)",
    range = c(1.5, 5)
  )+
  labs() +
  coord_cartesian(ylim = c(0,25)) + 
Theme() +
  theme(legend.position = "bottom",   axis.title.y = element_text(hjust = .9)
)
p
ggsave(
  filename = "figure_2.pdf",
  plot = p,
  device = cairo_pdf,
  width = 8.5,
  height = 5.5,
  units = "in",
  path = fig_path
)

ggsave(
  filename = "figure_2.png",
  plot = p + Theme(base_size = 32),
  device = ragg::agg_png,
  width = 8,
  height = 5,
  units = "in",
  dpi = 300,
  path = fig_path
)
median(campaign_df$duration)
quantile(campaign_df$duration, .25)
quantile(campaign_df$duration, .75)
campaign_df %>%
  group_by(period) %>%
  summarise(median = median(duration), IQR25 = quantile(duration, .25),  IQR = quantile(duration, .75))
campaign_df %>%
  arrange(rate_weighted)
```

```{r}
pub_table <- campaign_df %>%
  arrange((start)) %>%
  filter(!is.na(rate_weighted)) %>%
  transmute(
    `Start date` = start,
    `End date` = end, # force integer -> character (no commas)
    Country = country,
    `Target Population` = round(target_pop,0),
    `Coverage (%)` = round(coverage_weighted*100,2),
    `Vaccination rate` = round(rate_weighted*100,2)
  ) 

pub_table = pub_table %>%
  relocate(Country, .before = `Start date`)

# Build display table (convert numerics to plain character, no comma formatting)
# display_tbl <- pub_table %>%
#   mutate(
#     `Days: Index → Reactive Vaccination` = as.character(`Days: Index → Reactive Vaccination`),
#     `Total Cases` = ifelse(is.na(`Total Cases`), "Unknown", as.character(as.integer(`Total Cases`))),
#     `Total Deaths` = ifelse(is.na(`Total Deaths`), "Unknown", as.character(as.integer(`Total Deaths`))),
#     `Confirmed Cases` = ifelse(is.na(`Confirmed Cases`), "Unknown", as.character(as.integer(`Confirmed Cases`))),
#     `Confirmed Deaths` = ifelse(is.na(`Confirmed Deaths`), "Unknown", as.character(as.integer(`Confirmed Deaths`)))
#   )

ft_pub <- flextable(pub_table) %>%
  theme_booktabs() %>%
  bold(part = "header") %>%
  align(j = 1, align = "left", part = "body") %>%
  align(j = 2:ncol(pub_table), align = "center", part = "all") %>%
  autofit() %>%
  fontsize(size = 10, part = "all") %>%
  set_header_labels(
    Country = "Country",
    `Start date` = "Start date",
    `End date` = "End date", 
    `Target Population` = "Target Population",
    `Coverage (%)` = "Coverage (%)",
    `Vaccination rate`  = "Vaccination rate (% per day)"
  )

# days_num_valid <- pub_table$`Days: Index → Reactive Vaccination`
# days_num_valid <- days_num_valid[!is.na(days_num_valid)]
# 
# footer_txt <- sprintf(
#   "Days: mean = %.1f; median = %.1f; mode = %s; IQR = %.1f–%.1f; range = %.1f–%.1f; n = %d complete outbreaks",
#   mean(days_num_valid),
#   median(days_num_valid),
#   names(sort(table(days_num_valid), decreasing = TRUE))[1],
#   quantile(days_num_valid, 0.25),
#   quantile(days_num_valid, 0.75),
#   min(days_num_valid),
#   max(days_num_valid),
#   length(days_num_valid)
# )

# ft_pub <- ft_pub %>%
#   add_footer_lines(values = footer_txt) %>%
#   align(align = "left", part = "footer")

# Save outputs
out_dir <- "~/Documents/Documents/Yellow Fever/YF/fig"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)
png_file <- file.path(out_dir, "table_3.png")
docx_file <- file.path(out_dir, "table_3.docx")

try(save_as_image(ft_pub, path = png_file, expand = 8, zoom = 2), silent = TRUE)
try(save_as_docx(ft_pub, path = docx_file), silent = TRUE)

ft_pub
```






